### 2019-06-28 paper.tex
initial commit

\documentclass[9pt,twocolumn,twoside,lineno]{gsag3jnl}

\articletype{inv} % article type
% {inv} Investigations
% {msr} Mutant Screen Reports
% {gs} Genomic Selection
% {goi} Genetics of Immunity 
% {gos} Genetics of Sex 
% {mp} Multiparental Populations

\title{Real Time Genome Scan Using GPU}

\author[$\ast$]{Xiaoqi Hu}
\author[$\ast$]{Saunak Sen}

\affil[$\ast$]{University of Tennessee Health Science Center}

\keywords{Linear Model \\ Genome Scan \\ GPU}

\runningtitle{G3 Journal Template on Overleaf} % For use in the footer

%% For the footnote.
%% Give the last name of the first author if only one author;
% \runningauthor{FirstAuthorLastname}
%% last names of both authors if there are two authors;
% \runningauthor{FirstAuthorLastname and SecondAuthorLastname}
%% last name of the first author followed by et al, if more than two authors.
\runningauthor{Hu \textit{et al.}}

\begin{abstract}
The abstract should be written for people who may not read the entire paper, so it must stand on its own. The impression it makes usually determines whether the reader will go on to read the article, so the abstract must be engaging, clear, and concise. In addition, the abstract may be the only part of the article that is indexed in databases, so it must accurately reflect the content of the article. A well-written abstract is the  most effective way to reach intended readers, leading to more robust search, retrieval, and usage of the article.

Please see additional guidelines notes on preparing your abstract below.
\end{abstract}

\begin{document}

\maketitle
\thispagestyle{firststyle}
\logomark
\articletypemark
\marginmark
\firstpagefootnote

% Use the \equalcontrib command to mark authors with equal
% contributions, using the relevant superscript numbers
\equalcontrib{1}
\equalcontrib{2}

\correspondingauthoraffiliation{3}{Corresponding author: Please insert the affiliation correspondence address and email for the corresponding author. The corresponding author should be marked with the relevant number in the author list, as shown in the example.}
\vspace{-34pt}% Only used for adjusting extra space in the left column of the first page

\noindent This G3: Genes|Genomes|Genetics journal template is provided to help you write your work in the correct journal format. Instructions for use are provided below. Note that by default line numbers are present to aid reviewers and editors in reading and commenting on your manuscript. To remove line numbers, remove the \texttt{lineno} option from the \verb|\documentclass| declaration.

\section{Guide to using this template in Overleaf}

This template is provided to help you prepare your article for submission to the \textit{Genetics}.

\section{Author Affiliations}

For the authors' names, indicate different affiliations with the symbols: $\ast$, $\dagger$, $\ddagger$, $\S$. After four authors, the symbols double, triple, quadruple, and so forth as required.

\section{Your Abstract}

In addition to the guidelines provided in the example abstract above, your abstract should:

\begin{itemize}
\item provide a synopsis of the entire article;
\item begin with the broad context of the study, followed by specific background for the study;
\item describe the purpose, methods and procedures, core findings and results, and conclusions of the study;
\item emphasize new or important aspects of the research;
\item engage the broad readership of G3 and be understandable to a diverse audience (avoid using jargon);
\item be a single paragraph of less than 250 words;
\item contain the full name of the organism studied;
\item NOT contain citations or abbreviations.
\end{itemize}

\section{Introduction}

For the introduction, authors should be mindful of the broad readership of the journal. The introduction should set the stage for the importance of the work to a generalist reader and draw the reader in to the specific study. The scope and impact of the work should be clearly stated.

In individual organisms where a mutant is being studied, the rationale for the study of that mutant must be clear to a geneticist not studying that particular organism. Similarly, study of particular phenotypes should be justified broadly and not on the basis of interest for that organism alone. General background on the importance of the genetic pathway and/or phenotype should be provided in a single, well-reasoned paragraph near the beginning of the introduction.

Authors are encouraged to:

\begin{itemize}
\item cite the supporting literature completely rather than select a subset of citations;
\item provide important background citations, including relevant review papers (to help orient the non-specialist reader);
\item to cite similar work in other organisms.
\end{itemize}

\section{Materials and Methods}
\label{sec:materials:methods}

Manuscripts submitted to G3 should contain a clear description of the experimental design in sufficient detail so that the experimental analysis could be repeated by another scientist. If the level of detail necessary to explain the protocol goes beyond two paragraphs, give a short description in the main body of the paper and prepare a detailed description for supporting information.  For example, details would include indicating how many individuals were used, and if applicable how individuals or groups were combined for analysis. If working with mutants indicate how many independent mutants were isolated. If working with populations indicate how samples were collected and whether they were random with respect to the target population.

\subsection{Statistical Analysis}

Indicate which statistical analysis has been performed and describe the method and model applied. If many genes were examined simultaneously, or many phenotypes, a multiple comparison correction should be used to control the type I error rate, or a rationale for not applying a correction must be provided. The type of correction applied should be clearly stated. It should also be clear whether the p-values reported are raw, or after correction. Corrected p-values are often appropriate, but raw p-values should be available in the supporting materials so that others may perform their own corrections.

\subsection{Data Availability}

At the end of the Materials and Methods section, include a statement on reagent and data availability. Please read the Data and Reagent Policy before writing the statement. Make sure to list the accession numbers or DOIs of any data you have placed in public repositories. List the file names and descriptions of any data you will upload as supplemental information. The statement should also include any applicable IRB numbers. You may include specifications for how to properly acknowledge or cite the data.

For example: Strains are available upon request. File S1 contains detailed descriptions of all supplemental files. File S2 contains SNP ID numbers and locations. File S3 contains genotypes for each individual. Sequence data are available at GenBank and the accession numbers are listed in File S3. Gene expression data are available at GEO with the accession number: GDS1234. Code used to generate the simulated data is provided in file S4.

\section{Results and Discussion}

The results and discussion should not be repetitive. The results section should give a factual presentation of the data and all tables and figures should be referenced; the discussion should not summarize the results but provide an interpretation of the results, and should clearly delineate between the findings of the particular study and the possible impact of those findings in a larger context. Authors are encouraged to cite recent work relevant to their interpretations. Present and discuss results only once, not in both the Results and Discussion sections. It is sometimes acceptable to combine results and discussion. The text should be as succinct as possible. Heed Strunk and White's dictum: "Omit needless words!"

\section{Additional guidelines}

\subsection{Numbers} In the text, write out numbers nine or less except as part of a date, a fraction or decimal, a percentage, or a unit of measurement. Use Arabic numbers for those larger than nine, except as the first word of a sentence; however, try to avoid starting a sentence with such a number.

\subsection{Units} Use abbreviations of the customary units of measurement only when they are preceded by a number: "3 min" but "several minutes". Write "percent" as one word, except when used with a number: "several percent" but "75\%." To indicate temperature in centigrade, use ° (for example, 37°); include a letter after the degree symbol only when some other scale is intended (for example, 45°K).

\subsection{Nomenclature and Italicization} Italicize names of organisms even when  when the species is not indicated.  Italicize the first three letters of the names of restriction enzyme cleavage sites, as in HindIII. Write the names of strains in roman except when incorporating specific genotypic designations. Italicize genotype names and symbols, including all components of alleles, but not when the name of a gene is the same as the name of an enzyme. Do not use "+" to indicate wild type. Carefully distinguish between genotype (italicized) and phenotype (not italicized) in both the writing and the symbolism.

\subsection{Cross References}
Use the \verb|\nameref| command with the \verb|\label| command to insert cross-references to section headings. For example, a \verb|\label| has been defined in the section \nameref{sec:materials:methods}.

\section{In-text Citations}

Add citations using the \verb|\citep{}| command, for example \citep{neher2013genealogies} or for multiple citations, \citep{neher2013genealogies, rodelsperger2014characterization}.

For examples of different references, please see the example bibliography file (accessible via the Project menu in the Overleaf editor). This contains examples of articles \citep{neher2013genealogies, rodelsperger2014characterization}, a book \citep{Sturtevent2001}, a book chapter \citep{Sturtevent2001chp7}, ahead-of-print work \citep{Starita2015}, a preprint \citep{Falush16} and software \citep{Kruijer2015}.

\section{Examples of Article Components}
\label{sec:examples}

The sections below show examples of different header levels, which you can use in the primary sections of the manuscript (Results, Discussion, etc.) to organize your content.

\section{First level section header}

Use this level to group two or more closely related headings in a long article.

\subsection{Second level section header}

Second level section text.

\subsubsection{Third level section header:}

Third level section text. These headings may be numbered, but only when the numbers must be cited in the text.

\section{Figures and Tables}

Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.

\subsection{Sample Figure}

Figure \ref{fig:spectrum} shows an example figure.

\begin{figure}[htbp]

\label{fig:spectrum}
\end{figure}

\begin{figure}[htbp]

\label{video:spectrum}
\end{figure}

\subsection{Sample Table}

Table \ref{tab:shape-functions} shows an examplex table. Avoid shading, color type, line drawings, graphics, or other illustrations within tables. Use tables for data only; present drawings, graphics, and illustrations as separate figures. Histograms should not be used to present data that can be captured easily in text or small tables, as they take up much more space.

Tables numbers are given in Arabic numerals. Tables should not be numbered 1A, 1B, etc., but if necessary, interior parts of the table can be labeled A, B, etc. for easy reference in the text.

\begin{table*}[htbp]
\renewcommand{\familydefault}{\sfdefault}\normalfont
\centering
\caption{\bf Students and their grades}
\begin{tableminipage}{\textwidth}
\begin{tabularx}{\textwidth}{XXXX}
\hline
\header Student & Grade\footnote{This is an example of a footnote in a table. Lowercase, superscript italic letters (a, b, c, etc.) are used by default. You can also use *, **, and *** to indicate conventional levels of statistical significance, explained below the table.} & Rank & Notes \\
\hline
Alice & 82\% & 1 & Performed very well.\\
Bob & 65\% & 3 & Not up to his usual standard.\\
Charlie & 73\% & 2 & A good attempt.\\
\hline
\end{tabularx}
  \label{tab:shape-functions}
\end{tableminipage}
\end{table*}

\section{Sample Equation}

Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
\begin{equation}
S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
      = \frac{1}{n}\sum_{i}^{n} X_i
\label{eq:refname1}
\end{equation}
denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.

\bibliography{example-bibliography}

\end{document}

### 2019-07-03 paper.tex
moving poster content over

@@
 \author[$\ast$]{Xiaoqi Hu}
+\author[$\ast$]{Hyeonju Kim}
 \author[$\ast$]{Saunak Sen}
@@
 \begin{abstract}
-The abstract should be written for people who may not read the entire paper, so it must stand on its own.
-The impression it makes usually determines whether the reader will go on to read the article, so the abstract must be engaging, clear, and concise.
-In addition, the abstract may be the only part of the article that is indexed in databases, so it must accurately reflect the content of the article.
-A well-written abstract is the  most effective way to reach intended readers, leading to more robust search, retrieval, and usage of the article.
+The BXD strains are an important reference population of
+recombinant inbred lines which have been phenotyped extensively.
+To facilitate interactive use of genotype-phenotype relationships
+in this population, we sought to speed up eQTL scans where we
+perform a univariate genome scan for every trait in a collection
+of omic traits.  By using easily parallelizable operations such as
+matrix multiplication, vectorized operations, and elementwise
+operations, we are able to decrease runtimes approaching real-time
+computation.  We used parallelization using different CPU threads
+as well as GPUs.  We found that the speed advantage of GPUs is
+dependent on problem size and shape.  These results indicate a pathway for
+speeding up eQTL scans using LMMs.  Our implementation is in the
+Julia programming language.
 
-Please see additional guidelines notes on preparing your abstract below.
 \end{abstract}
@@
 
-\noindent This G3: Genes|Genomes|Genetics journal template is provided to help you write your work in the correct journal format.
-Instructions for use are provided below.
-Note that by default line numbers are present to aid reviewers and editors in reading and commenting on your manuscript.
-To remove line numbers, remove the \texttt{lineno} option from the \verb|\documentclass| declaration.
+\noindent        Computational demands in omics are growing exponentially. For
+instance, eQTL analysis requires processing genomewide
+expression and marker data; this can take minutes, hours or
+days depending on problem size.  Scientists want to intereact
+with the results and real-time processing would be valuable.
+Our goal is to build a backend for GeneNetwork that will enable
+real-time genome scans for important genetic reference
+populations such as the BXD population.
 
-\section{Guide to using this template in Overleaf}
+\section{Methods and Data}
+\label{sec:methods:Data}
 
-This template is provided to help you prepare your article for submission to the \textit{Genetics}.
-
-\section{Author Affiliations}
-
-For the authors' names, indicate different affiliations with the symbols: $\ast$, $\dagger$, $\ddagger$, $\S$. After four authors, the symbols double, triple, quadruple, and so forth as required.
-
-\section{Your Abstract}
-
-In addition to the guidelines provided in the example abstract above, your abstract should:
-
-\begin{itemize}
-\item provide a synopsis of the entire article;
-\item begin with the broad context of the study, followed by specific background for the study;
-\item describe the purpose, methods and procedures, core findings and results, and conclusions of the study;
-\item emphasize new or important aspects of the research;
-\item engage the broad readership of G3 and be understandable to a diverse audience (avoid using jargon);
-\item be a single paragraph of less than 250 words;
-\item contain the full name of the organism studied;
-\item NOT contain citations or abbreviations.
-\end{itemize}
-
-\section{Introduction}
-
-For the introduction, authors should be mindful of the broad readership of the journal.
-The introduction should set the stage for the importance of the work to a generalist reader and draw the reader in to the specific study.
-The scope and impact of the work should be clearly stated.
-
-In individual organisms where a mutant is being studied, the rationale for the study of that mutant must be clear to a geneticist not studying that particular organism.
-Similarly, study of particular phenotypes should be justified broadly and not on the basis of interest for that organism alone.
-General background on the importance of the genetic pathway and/or phenotype should be provided in a single, well-reasoned paragraph near the beginning of the introduction.
-
-Authors are encouraged to:
-
-\begin{itemize}
-\item cite the supporting literature completely rather than select a subset of citations;
-\item provide important background citations, including relevant review papers (to help orient the non-specialist reader);
-\item to cite similar work in other organisms.
-\end{itemize}
-
-\section{Materials and Methods}
-\label{sec:materials:methods}
-
-Manuscripts submitted to G3 should contain a clear description of the experimental design in sufficient detail so that the experimental analysis could be repeated by another scientist.
+%Manuscripts submitted to G3 should contain a clear description of the experimental design in sufficient detail so that the experimental analysis could be repeated by another scientist.
 If the level of detail necessary to explain the protocol goes beyond two paragraphs, give a short description in the main body of the paper and prepare a detailed description for supporting information.
@@
 
-\subsection{Statistical Analysis}
+\subsection{Linear Model} 
+ Let $y_i$ denote a vector of gene expressions for $n$
+individuals in the $i$-th omics trait ( $i=1,\ldots,m$).  We
+define a univariate linear model as follows:
 
-Indicate which statistical analysis has been performed and describe the method and model applied.
+\begin{eqnarray*}
+	y_i &=& X_j \beta_j+ \epsilon_i,
+	\quad \epsilon_i \sim N(0,\sigma_i^2)
+\end{eqnarray*}
+
+where ${X}_j$ is a matrix that may include the $y-$intercept with
+or without covariate(s), and the $j$-th candidate genetic marker
+($j=1,\ldots,p$).  ${\beta}_j$ is a vector of the $j$-th eQTL
+effects, and ${\epsilon}_i$ is random error.  Suppose $RSS_{0i}$
+is a residual sum of squares under the null hypothesis of no eQTL
+and $RSS_{1ij}$ is the residual sum of squares under the
+alternative for the $i$-th trait and $j$-th marker.  Then the LOD
+score for the corresponding pair can be written as:
+
+\begin{eqnarray*}
+	LOD_{ij} &=& \frac{n}{2} \log_{10} \left( \frac{RSS_{0i}}{RSS_{1ij}} \right)\\
+	&=& \frac{n}{2} \log_{10} \left( 1{-}r_{ij}^2 \right),
+\end{eqnarray*}
+where $r_{ij}$ is the correlation between the $i$-th trait and
+$j$-th marker. Note that this only works for one-df tests.  If the
+trait matrix and genotype matrix ($G$) are standardized, then the
+correlation matrix is simply $$R=Y^{'}G.$$
+%Indicate which statistical analysis has been performed and describe the method and model applied.
 If many genes were examined simultaneously, or many phenotypes, a multiple comparison correction should be used to control the type I error rate, or a rationale for not applying a correction must be provided.
@@
 
-\subsection{Data Availability}
+\subsection{Acceleration Techniques}
+We used various techniques to accelerate eQTL scans. 
+\subsubsection{Multithreaded CPU operations}
+\subsubsection{Matrix and Vectorized Operations}
+\subsubsection{Single precision}
+\subsubsection{GPU operations}
+\subsubsection{Julia language}
 
-At the end of the Materials and Methods section, include a statement on reagent and data availability.
+\subsection{Information and Availability of Datasets}
+Data from 79 BXD strains for 35556 transcripts from spleens were collected using the Affy Mouse Gene 1.0 ST array.
+Animals were generated by Drs.
+Lu, Williams, and colleagues.
+There were usu- ally one male and one female per strain; strain averages were used as the trait of analysis.
+All animals were young adults (oldest 160 days).
+GN accession: GN283.
+Marker data was available on 7321 markers; genotype probabilities were calculated using R/qtl.
+
+      \begin{figure}[!htb]
+
+	\caption{Schematic of data and correlation calculation: $Y$ is
+		the expression phenotype matrix, $G$ is the matrix of
+		genotytpes, and $R=Y^{\prime}G$ is the matrix of
+		correlations.  The LOD scores are a function of the
+		correlation matrix.}
+	\label{MatrixMult}
+	\includegraphics[scale = .4]{figs/YGmatrix.png}
+\end{figure}
+
+%At the end of the Materials and Methods section, include a statement on reagent and data availability.
 Please read the Data and Reagent Policy before writing the statement.
@@
 
-For example: Strains are available upon request.
+%For example: Strains are available upon request.
 File S1 contains detailed descriptions of all supplemental files.
@@
 
-\section{Results and Discussion}
+\section{Results}
+{\em Profiling:} Calculating correlation matrix and LOD score
+takes up to 90\% of runtime.  Both involve matrix operations
+(multiplication and elementwise operations respectively) which
+are highly parallelizable.
+      \begin{itemize}
+	\item We chose OpenBLAS library for matrix multiplication, an
+	optimized multithreaded library.  In our experiment, gslBLAS
+	(GNU Scientific Library) is 10 times slower than OpenBLAS.
+	\item We used multithreading operations whenever possible, such as
+	for elementwise operations.
+	\item We explored using single precision instead of double
+	precision.  Precision change affects both throughput (FLOPS) and
+	data transfer time.  Single precision is twice
+	as fast as double precision, and provides correct result
+	within 1e-2 tolerance (Table 1).
+\end{itemize}
 
-The results and discussion should not be repetitive.
+ {\em GPUs vs CPUs:} GPUs are attractive since they provide a
+massive number of cores at a lower price range.
+Matrix multiplication and element-wise operations are amendable
+to GPU's heterogeneous computing architecture, since both have
+no data race conditions and low data dependencies.
+
+\begin{itemize}
+	\item Matrix multiplication is up to 5 times faster
+	compared to 16-threaded CPU.
+	% We compared the speed of matrix
+	% multiplication on CPU and GPU, with matrix sizes
+	% ranging from 1k to 100k.
+	We found speed
+	up is sensitive to the size and the shape of matrices.
+	% GPU is best used for large number of operations, therefore
+	% multiplying larger matrices will give better speed up.  The
+	% shape of matrices also plays an important role when the matrix
+	% size is large enough.
+	Figure 2 shows the speedup of matrix
+	multiplication, when the total data size of input and output
+	floating point matrices is over 12 GB.
+	% Our reasoning for such
+	% sensitivity is that the data transfer from GPU(device) to
+	% CPU(host) is half of CPU to GPU.
+	When output data size is
+	larger than input matrix, the speed up will decrease. 
+	\item  GPU profiling shows that 98\% of eQTL scan for BXD
+	spleen dataset is spent on data transfer.
+	Minimizing data transfer time will make the scan significantly
+	faster.
+\end{itemize}
+
+      {\em Julia as development platform:} In our experiment
+of matrix multiplication, Julia's speed is comparable to C/C++.
+However, the low learning curve, clean syntax, as well as
+support to GPU programming libraries such as CUDAnative and
+CuArrays affords much lower programming efforts compared to
+C/C++.
+
+\begin{figure}[!htb]
+	\centering
+	\caption{Variation of GPU vs CPU speedup with matrix shape
+	}
+	\label{GPUCPUShape}
+	\includegraphics[scale = 0.4]{figs/speedup.png}
+\end{figure}
+
+\subsection{Platform}
+ \begin{itemize}
+	\item CPU: Intel Xeon Gold 6148; 40 cores @ 2.40GHz, 192GB 
+	\item GPU: Tesla V100-PCIE; 5120 Cores @ 1380 MHz, 16GB
+\end{itemize}
+Software: 
+\begin{itemize}
+	\item Programming environment: Julia; CentOS 7
+	\item Libraries: CUDA v10.1 and cuBLAS; OpenBLAS
+	\item Profilers: Julia Profiler Package; nvprof
+\end{itemize}
+
+%The results and discussion should not be repetitive.
 The results section should give a factual presentation of the data and all tables and figures should be referenced; the discussion should not summarize the results but provide an interpretation of the results, and should clearly delineate between the findings of the particular study and the possible impact of those findings in a larger context.
@@
 
-\section{Additional guidelines}
-
-\subsection{Numbers} In the text, write out numbers nine or less except as part of a date, a fraction or decimal, a percentage, or a unit of measurement.
-Use Arabic numbers for those larger than nine, except as the first word of a sentence; however, try to avoid starting a sentence with such a number.
-
-\subsection{Units} Use abbreviations of the customary units of measurement only when they are preceded by a number: "3 min" but "several minutes".
-Write "percent" as one word, except when used with a number: "several percent" but "75\%." To indicate temperature in centigrade, use ° (for example, 37°); include a letter after the degree symbol only when some other scale is intended (for example, 45°K).
-
-\subsection{Nomenclature and Italicization} Italicize names of organisms even when  when the species is not indicated.
-Italicize the first three letters of the names of restriction enzyme cleavage sites, as in HindIII.
-Write the names of strains in roman except when incorporating specific genotypic designations.
-Italicize genotype names and symbols, including all components of alleles, but not when the name of a gene is the same as the name of an enzyme.
-Do not use "+" to indicate wild type.
-Carefully distinguish between genotype (italicized) and phenotype (not italicized) in both the writing and the symbolism.
-
-\subsection{Cross References}
-Use the \verb|\nameref| command with the \verb|\label| command to insert cross-references to section headings.
-For example, a \verb|\label| has been defined in the section \nameref{sec:materials:methods}.
-
-\section{In-text Citations}
-
-Add citations using the \verb|\citep{}| command, for example \citep{neher2013genealogies} or for multiple citations, \citep{neher2013genealogies, rodelsperger2014characterization}.
-
-For examples of different references, please see the example bibliography file (accessible via the Project menu in the Overleaf editor).
-This contains examples of articles \citep{neher2013genealogies, rodelsperger2014characterization}, a book \citep{Sturtevent2001}, a book chapter \citep{Sturtevent2001chp7}, ahead-of-print work \citep{Starita2015}, a preprint \citep{Falush16} and software \citep{Kruijer2015}.
-
-\section{Examples of Article Components}
-\label{sec:examples}
-
-The sections below show examples of different header levels, which you can use in the primary sections of the manuscript (Results, Discussion, etc.) to organize your content.
-
-\section{First level section header}
-
-Use this level to group two or more closely related headings in a long article.
-
-\subsection{Second level section header}
-
-Second level section text.
-
-\subsubsection{Third level section header:}
-
-Third level section text. These headings may be numbered, but only when the numbers must be cited in the text.
-
-\section{Figures and Tables}
-
-Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.
-
-\subsection{Sample Figure}
-
-Figure \ref{fig:spectrum} shows an example figure.
-
-\begin{figure}[htbp]
-
-\label{fig:spectrum}
-\end{figure}
-
-\begin{figure}[htbp]
-
-\label{video:spectrum}
-\end{figure}
+\section{Conclusion}
 
@@
 Table \ref{tab:shape-functions} shows an examplex table.
-Avoid shading, color type, line drawings, graphics, or other illustrations within tables.
-Use tables for data only; present drawings, graphics, and illustrations as separate figures.
-Histograms should not be used to present data that can be captured easily in text or small tables, as they take up much more space.
-
-Tables numbers are given in Arabic numerals. Tables should not be numbered 1A, 1B, etc., but if necessary, interior parts of the table can be labeled A, B, etc. for easy reference in the text.
 
@@
 
-\section{Sample Equation}
-
-Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
-\begin{equation}
-S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
-      = \frac{1}{n}\sum_{i}^{n} X_i
-\label{eq:refname1}
-\end{equation}
-denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.
-
 \bibliography{example-bibliography}

### 2019-07-03 paper.tex
adding in details in *acceleration techniques* part

@@
 \subsubsection{Multithreaded CPU operations}
+ Used for multi-threaded loops whenever possible, eg. for element-wise operations
 \subsubsection{Matrix and Vectorized Operations}
+ OpenBLAS for multi-threaded matrix operations. 
+Matrix multiplication is a well studied area where the computer scientists try to apply various optimization techniques to make it run faster. 
+There are various BLAS libraries available, such as gslBLAS, OpenBLAS, etc. 
+The matrix multiplication in OpenBLAS is default multithreaded, therefore does not require extra coding effort to achieve highly efficient result.
+
 \subsubsection{Single precision}
+ Faster and memory efficient; may be less accurate, but acceptable for quick scan
+Most dataset used in scientific computing are stored in double precision.
+
 \subsubsection{GPU operations}
+Specialized computations such as matrix multiplications and element-wise operations; requires additional programming effort
 \subsubsection{Julia language}
+Programming simplicity of interpreted languages with speed approaching compiled languages; GPU programming and linear algebra straightforward
+\subsubsection{ Heritability grid}
+For LMMs, considering a discrete grid of heritabilities enhances parallelization
 

### 2019-07-08 paper.tex
adding cpu and gpu time comparison table

--- 2019-07-03 paper.tex
@@
 % {mp} Multiparental Populations
+\usepackage{verbatim}
+\usepackage{multirow}
 
@@
 \subsubsection{Multithreaded CPU operations}
+ Used for multi-threaded loops whenever possible, eg. for element-wise operations
 \subsubsection{Matrix and Vectorized Operations}
+% OpenBLAS for multi-threaded matrix operations. 
+Matrix multiplication is a well studied area where the computer scientists try to apply various optimization techniques to make it run faster. 
+There are various BLAS libraries available, such as gslBLAS, OpenBLAS, etc. 
+The matrix multiplication in OpenBLAS is default multithreaded, therefore does not require extra coding effort to achieve highly efficient result.
+
 \subsubsection{Single precision}
+% Faster and memory efficient; may be less accurate, but acceptable for quick scan
+Precision, in computer science, means the smallest difference between two representable numbers.
+Floating point numbers, in scientific computing, are usually stored in double presicision. 
+Double precision floating point numebrs takes up 8 bytes in memory, while single precision takes up 4 bytes. 
+In addition to the difference in storing size, the speed for calculation using single and double precision also varies depending on hardware. 
+For example, the GPU throughput (number of floating point calculation per second, aka FLOPS) for double precision is 1/32 of single precision on a Nvidia GTX 1050 GPU, and 1/4 on a Nvidia Tesla K80. 
+Differences in optimization techniques, underlying architecture cause such performance gap. 
+Considering the memory size, data transfer speed and FLOPS, single precision brings multiple benefits when precision is not the priority concern.
+
 \subsubsection{GPU operations}
+\iffalse
+% this block is commented out 
+	In order to achieve the goal of real time eQTL scans, we profiled the entire process of eQTL scan.
+Profiling can provide insightful information about the program,  such as running time of each function.
+	Targeting the longest running time functions provides the most gain with limited programming effort and computing resources. 
+	In our profiling result, we found that the is calculating correlation matrix and LOD score takes up to 90\% of total running time. 
+	Since both steps uses matrix multiplication, which is highly parallelizable and well-researched, we choose the following approaches to speed up matrix multiplication. 
+	First, we chose OpenBLAS library for matrix multiplication, which is an optimized  multithreaded BLAS library. 
+	In our experiment, gslBLAS (GNU Scientific Library) is 10 times slower than OpenBLAS. 
+	In addition to choosing an optimized multithreaded library, we use multithreading operations whenver possible.
+	Second, we considered the possibility of using single precision compared to double precision. 
+	Precision change will affect both throughput and data transfer time. 
+	Throughput is the number of operations done in unit time, usually measured in FLOPS(floating point operations).
+	In our experiment, single precision is twice as fast as double precision, and provides correct result within 1e-2 tolerance. 
+	Detailed running time of single precision and double precision is show in Table 1. 
+	Third, we considered GPU as an attractive computing platform, because of its massive number of cores at a lower price range, and fast open source GPGPU libraries such as CUDA and OpenCL. 
+	Matrix multiplication and element-wise operations are amendable to GPU's heterogeneous computing architecture, since both have no data race conditions and low data dependencies. 
+	In our experiment, matrix multiplication is up to 5 times faster compared to 16 threaded CPU. 
+	We compared the speed of matrix multiplication on CPU and GPU, with rectangular matrix sizes ranging from 1k to 100k, increased by power of 2, 
+	We found speed up is sensitive to the size of matrix, and the shape of matrix.  
+	GPU is best used for large number of operations, therefore multiplying larger matrices will give better speed up. 
+	The shape of matrices also plays an important role when the matrix size is large enough. 
+	Figure 2 shows the speed up of matrix multiplication, when the total data size of input and output floating point matrices is over 12 GB. 
+	Our reasoning for such sensitivity is that the data transfer from GPU(device) to CPU(host) is half of CPU to GPU. 
+	When output data size is larger than input matrix, the speed up will decrease. 
+	Further experiment will be done to prove such hypothesis. 
+	In addition, our GPU profiling result shows that 98\% of eQTL scan for BXD spleen dataset is spent on data transfer for the GPU approach. 
+	Minimizing data transfer time will make the scan significantly faster. 
+	Further programming efforts will make GPU a more attractive approach to reach real time eQTL scans. 
+	Lastly but most fundamentally, we choose Julia programming language as our development tool. 
+	In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
+	However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative and CuArrays affords much lower programming efforts compared to C/C++.
+\fi
+
+%Specialized computations such as matrix multiplications and element-wise operations; requires additional programming effort
+Originally used in graphics, GPU has taken off as a general computing device in recent years because of its massive number of cores at a lower price range, and fast GPGPU libraries such as CUDA and OpenCL.
+Based on our profiling result, the time consuming parts of our genome scan method are matrix multiplication and element-wise operations. 
+Both are amendable to GPU's heterogeneous computing architecture, since they have no data race conditions and low data dependencies. 
+However, GPU also has its own limitations. 
+To truely utilize the maximum computing power of GPU, one needs to think creatively to work around such limitations from GPU. 
+For example, during our experiemnt, memory transfer between host and device is really slow. 
+Profiling result shows that 98\% of total genome scan time is spent on memory transfer. 
+To cope with this limitation, instead of off-loading the entire correlation matrix, we use GPU to calculate the max lod score of each phenotype, and output the max.
+
 \subsubsection{Julia language}
+%Programming simplicity of interpreted languages with speed approaching compiled languages; GPU programming and linear algebra straightforward
+In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
+However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative and CuArrays affords much lower programming efforts compared to C/C++.
+%TODO: Add reference to CUDAnative and CuArrays
+\subsubsection{Heritability grid}
+For LMMs, considering a discrete grid of heritabilities enhances parallelization
 
@@
 Lu, Williams, and colleagues.
-There were usu- ally one male and one female per strain; strain averages were used as the trait of analysis.
+There were usually one male and one female per strain; strain averages were used as the trait of analysis.
 All animals were young adults (oldest 160 days).
@@
 \section{Results}
+We profiled the entire genome scan program to get a complete picture of the execution process.
+Our profiling result shows that 90\% of runtime is spent on calculating correlation matrix and LOD score. 
+Since both involve matrix operations (multiplication and elementwise operations respectively) which are highly parallelizable, we used GPU to speed up matrix operations.
+
+Table \ref{tab:speedup-table} shows timing between CPU and GPU
+
+\begin{table*}[htbp]
+	\renewcommand{\familydefault}{\sfdefault}\normalfont
+	\centering
+	\caption{\bf Time comparison between CPU and GPU}
+	\begin{tableminipage}{\textwidth}
+		\begin{tabularx}{\textwidth}{XXXX}
+			\hline
+			\header Method & Precision & CPU   & GPU  \\
+			\hline
+			\multirow{2}{*}{No Find Max} & Single    & 0.36s & 0.41s \\
+			& Double    & 0.55s & 0.80s \\
+			\multirow{2}{*}{Find Max}    & Single    &       &       \\
+			& Double    & 1.77s & 0.08s \\
+			\hline
+
+		\end{tabularx}
+		\label{tab:speedup-table}
+	\end{tableminipage}
+\end{table*}
+
 {\em Profiling:} Calculating correlation matrix and LOD score
@@
 	\label{GPUCPUShape}
-	\includegraphics[scale = 0.4]{figs/speedup.png}
+	\includegraphics[scale = 0.36]{figs/speedup.png}
 \end{figure}
@@
 
-\subsection{Sample Table}
-
-Table \ref{tab:shape-functions} shows an examplex table.
-
-\begin{table*}[htbp]
-\renewcommand{\familydefault}{\sfdefault}\normalfont
-\centering
-\caption{\bf Students and their grades}
-\begin{tableminipage}{\textwidth}
-\begin{tabularx}{\textwidth}{XXXX}
-\hline
-\header Student & Grade\footnote{This is an example of a footnote in a table.
-Lowercase, superscript italic letters (a, b, c, etc.) are used by default.
-You can also use *, **, and *** to indicate conventional levels of statistical significance, explained below the table.} & Rank & Notes \\
-\hline
-Alice & 82\% & 1 & Performed very well.\\
-Bob & 65\% & 3 & Not up to his usual standard.\\
-Charlie & 73\% & 2 & A good attempt.\\
-\hline
-\end{tabularx}
-  \label{tab:shape-functions}
-\end{tableminipage}
-\end{table*}
-
 \bibliography{example-bibliography}

### 2019-07-09 paper.tex
adding in timing for single precision using find max

@@
 Corrected p-values are often appropriate, but raw p-values should be available in the supporting materials so that others may perform their own corrections.
+Figure \ref{MatrixMult} shows the data layout.
+      \begin{figure}[!htb]
+
+	\caption{Schematic of data and correlation calculation: $Y$ is
+		the expression phenotype matrix, $G$ is the matrix of
+		genotytpes, and $R=Y^{\prime}G$ is the matrix of
+		correlations.  The LOD scores are a function of the
+		correlation matrix.}
+	\label{MatrixMult}
+	\includegraphics[scale = .4]{figs/YGmatrix.png}
+\end{figure}
 
@@
 In addition to the difference in storing size, the speed for calculation using single and double precision also varies depending on hardware. 
-For example, the GPU throughput (number of floating point calculation per second, aka FLOPS) for double precision is 1/32 of single precision on a Nvidia GTX 1050 GPU, and 1/4 on a Nvidia Tesla K80. 
+For example, the GPU throughput (number of floating point calculation per second, measured in FLOPS) for double precision is 1/32 of single precision on a Nvidia GTX 1050 GPU, and 1/4 on a Nvidia Tesla K80.
 Differences in optimization techniques, underlying architecture cause such performance gap. 
-Considering the memory size, data transfer speed and FLOPS, single precision brings multiple benefits when precision is not the priority concern.
+Considering the memory size, data transfer speed and throughput, single precision brings multiple benefits when precision is not the priority concern.
 
@@
 
-      \begin{figure}[!htb]
-
-	\caption{Schematic of data and correlation calculation: $Y$ is
-		the expression phenotype matrix, $G$ is the matrix of
-		genotytpes, and $R=Y^{\prime}G$ is the matrix of
-		correlations.  The LOD scores are a function of the
-		correlation matrix.}
-	\label{MatrixMult}
-	\includegraphics[scale = .4]{figs/YGmatrix.png}
-\end{figure}
-
-%At the end of the Materials and Methods section, include a statement on reagent and data availability.
-Please read the Data and Reagent Policy before writing the statement.
-Make sure to list the accession numbers or DOIs of any data you have placed in public repositories.
-List the file names and descriptions of any data you will upload as supplemental information.
-The statement should also include any applicable IRB numbers.
-You may include specifications for how to properly acknowledge or cite the data.
-
-%For example: Strains are available upon request.
-File S1 contains detailed descriptions of all supplemental files.
-File S2 contains SNP ID numbers and locations.
-File S3 contains genotypes for each individual.
-Sequence data are available at GenBank and the accession numbers are listed in File S3.
-Gene expression data are available at GEO with the accession number: GDS1234.
-Code used to generate the simulated data is provided in file S4.
-
 \section{Results}
@@
 			& Double    & 0.55s & 0.80s \\
-			\multirow{2}{*}{Find Max}    & Single    &       &       \\
-			& Double    & 1.77s & 0.08s \\
+			\multirow{2}{*}{Find Max}    & Single    &1.54s       &0.056s       \\
+			& Double    & 1.77s & 0.079s \\
 			\hline
@@
 no data race conditions and low data dependencies.
+\begin{figure}[!htb]
+	\centering
+	\caption{Variation of GPU vs CPU speedup with matrix shape
+	}
+	\label{GPUCPUShape}
+	\includegraphics[scale = 0.36]{figs/speedup.png}
+\end{figure}
 
@@
 
-\begin{figure}[!htb]
-	\centering
-	\caption{Variation of GPU vs CPU speedup with matrix shape
-	}
-	\label{GPUCPUShape}
-	\includegraphics[scale = 0.36]{figs/speedup.png}
-\end{figure}
-
 \subsection{Platform}

### 2019-07-09 paper.tex
writing result section, before writing GPU part

@@
 \section{Results}
+\subsection{Benefit of using linear model}
+Our method uses linear model, and simplified the genome scan process to basic matrix operations. 
+The timing of our method is shown in Table \ref{tab:speedup-table}. 
+The execution time without finding the maximum LOD score of phenotype using double precision is 0.55 second if we only use CPU. 
+Getting the same result using the same dataset from R/qtl took about 36.5 seconds.
+
+\subsection{Benefit of using single precision}
+Table \ref{tab:speedup-table} also shows the execution time using single and double precisions. 
+In all cases, genome scan using single precision runs faster than double precision. 
+Using single precision provides benefits in 3 aspects: memory storage, data transfer, and arithmetic calculation.
+
+\subsection{Benefit of using GPU}
+To make fair comparison for CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
+It includes the following steps (step 4 is optional):
+\begin{itemize}
+	\item Calculate standadized matrix for input matrices.
+	\item Get correlation matrix
+	\item Caclulate LOD score
+	\item Find maximum LOD score for each phenotype. 
+\end{itemize}
+
 We profiled the entire genome scan program to get a complete picture of the execution process.
@@
 			\hline
-			\header Method & Precision & CPU   & GPU  \\
+			\header Method & Precision & CPU only   & CPU \& GPU  \\
 			\hline
-			\multirow{2}{*}{No Find Max} & Single    & 0.36s & 0.41s \\
+			\multirow{2}{*}{No Find Max LOD} & Single    & 0.36s & 0.41s \\
 			& Double    & 0.55s & 0.80s \\
-			\multirow{2}{*}{Find Max}    & Single    &1.54s       &0.056s       \\
+			\multirow{2}{*}{Find Max LOD}    & Single    &1.54s       &0.056s       \\
 			& Double    & 1.77s & 0.079s \\

### 2019-07-10 paper.tex
adding GPU part in result

@@
 
-\section{Results}
-\subsection{Benefit of using linear model}
-Our method uses linear model, and simplified the genome scan process to basic matrix operations. 
-The timing of our method is shown in Table \ref{tab:speedup-table}. 
-The execution time without finding the maximum LOD score of phenotype using double precision is 0.55 second if we only use CPU. 
-Getting the same result using the same dataset from R/qtl took about 36.5 seconds.
-
-\subsection{Benefit of using single precision}
-Table \ref{tab:speedup-table} also shows the execution time using single and double precisions. 
-In all cases, genome scan using single precision runs faster than double precision. 
-Using single precision provides benefits in 3 aspects: memory storage, data transfer, and arithmetic calculation.
-
-\subsection{Benefit of using GPU}
-To make fair comparison for CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
-It includes the following steps (step 4 is optional):
-\begin{itemize}
-	\item Calculate standadized matrix for input matrices.
-	\item Get correlation matrix
-	\item Caclulate LOD score
-	\item Find maximum LOD score for each phenotype. 
-\end{itemize}
-
-We profiled the entire genome scan program to get a complete picture of the execution process.
-Our profiling result shows that 90\% of runtime is spent on calculating correlation matrix and LOD score. 
-Since both involve matrix operations (multiplication and elementwise operations respectively) which are highly parallelizable, we used GPU to speed up matrix operations.
-
-Table \ref{tab:speedup-table} shows timing between CPU and GPU
-
 \begin{table*}[htbp]
@@
 		\end{tabularx}
+		\footnotetext[1]{There are overhead of using GPU, such as data transfer and setting up context, the timing shown for CPU \& GPU here included all overhead for fairness.
+We ran the genome scan process 10 times and choose the median to remove randomness of each run, and the warm up time of GPU.}
 		\label{tab:speedup-table}
@@
 \end{table*}
+
+\section{Results}
+\subsection{Benefit of using linear model}
+Our method uses linear model, and simplified the genome scan process to basic matrix operations. 
+The timing of our method is shown in Table \ref{tab:speedup-table}. 
+The execution time without finding the maximum LOD score of phenotype using double precision is 0.55 second if we only use CPU. 
+Getting the same result using the same dataset from R/qtl took about 36.5 seconds.
+
+\subsection{Benefit of using single precision}
+Table \ref{tab:speedup-table} also shows the execution time using single and double precisions. 
+In all cases, genome scan using single precision runs faster than double precision. 
+Using single precision provides benefits in 3 aspects: memory storage, data transfer, and arithmetic calculation.
+
+\subsection{Benefit of using GPU}
+In parallel computing, Amdahl's law is used to find the theoretical maximum speedup by improving a particular part of a program.  
+For example, if a program took ten minutes for a serial processor, and a function that takes nine our of those ten minutes can be parallelized, then, the theoretical speedup, no matter how many processors are used, can not be more than ten times, because the minimum execution time of this program is one minute.
+Therefore, profiling the entire genome scan process is the prerequisit in order to begin the optimization. 
+Often, profiling would consider space and time complexity. 
+Our primary concern is the time taken by each function, therefore only timing information is considered in our profiling. 
+We used Julia's built-in Sampling Profiler (%TODO reference here.)
+to find our target function for GPU because it is less intrusive compared to other profiling methods.
+
+The genome scan process includes the following steps:
+\begin{itemize}
+	\item Calculate standadized matrix for input matrices.
+	\item Get correlation matrix
+	\item Caclulate LOD score
+\end{itemize}
+Our profiling result shows that step two and three takes up 90\% of time, and both involve parallelizable matrix operations. 
+So they are our candidates for GPU accleration.
+To make fair comparison for CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
+The \textit{No Find Max LOD} method shows the timing when running the scan with \textit{CPU only} and \textit{CPU\&GPU}.
+As the table shows, the CPU\&GPU combination did not show any improvement over the serial version (CPU only) in terms of time. 
+In contrast, it is slower, whether using single or double precision for the data type.
+We used a GPU profiler \textit{nvprof} %TODO reference here
+to investigate what is the bottleneck of GPU. 
+Result is, 98\% of time is spent on data transfer from GPU to CPU (device to host).
+As shown in Figure \ref{MatrixMult}, the input matrices Y' and G are small compared to the output matrix R.
+For BXD spleen dataset, the Y' and G matrix is about 40MB %TODO: double check on this number
+, while R matrix is about 4GB.
+For our test hardware Nvidia TESLA V100, data transfer rate from CPU to GPU (host to device) is 4.3GB/s, and GPU to CPU (device to host) is 2.2GB/s. %TODO: double check the numbers here. 
+Larger data size at a slower transfer rate makes GPU seems unattractive for computing genome scan, at least for BXD dataset.
+
+To cope with this limitaion of GPU, we further developed \textit{Find Max LOD} method, because main interest of genome scan is to find the maximum LOD score of every phenotype. 
+This step is highly parallelizable and can utilize GPU's massive cores, at the same time, reduce the amount of data needed to be transfered back to host. 
+Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
+Compared to CPU only, which took 1.77 seconds, this method gains 22 times speedup.
+
+\begin{figure}[!htb]
+	\centering
+	\caption{Variation of GPU vs CPU speedup with matrix shape
+	}
+	\label{GPUCPUShape}
+	\includegraphics[scale = 0.36]{figs/speedup.png}
+\end{figure}
+
+\iffalse
+We profiled the entire genome scan program to get a complete picture of the execution process.
+Our profiling result shows that 90\% of runtime is spent on calculating correlation matrix and LOD score. 
+Since both involve matrix operations (multiplication and elementwise operations respectively) which are highly parallelizable, we used GPU to speed up matrix operations.
 
@@
 no data race conditions and low data dependencies.
-\begin{figure}[!htb]
-	\centering
-	\caption{Variation of GPU vs CPU speedup with matrix shape
-	}
-	\label{GPUCPUShape}
-	\includegraphics[scale = 0.36]{figs/speedup.png}
-\end{figure}
 
@@
 
+\fi
+
 \subsection{Platform}

### 2019-07-10 paper.tex
adding affiliations

@@
 \author[$\ast$]{Hyeonju Kim}
+\author[$\ast$]{GregoryFarage}
+\author[$\dagger$]{PjotrPrins}
+\author[$\dagger$]{RobertW.Williams}
+\author[$\S$]{KarlW.Broman}
 \author[$\ast$]{Saunak Sen}
 
-\affil[$\ast$]{University of Tennessee Health Science Center}
+\affil[$\ast$]{Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis, TN} 
+\affil[$\dagger$]{Department of Genetics, Genomics, and Informatics, University of Tennessee Health Science Center, Memphis, TN} 
+\affil[$\S$]{Department of Biostatistics, University of Wisconsin-Madison, Madison, WI}
 

### 2019-07-18 paper.tex
modified timing in table with hippocampus data

@@
 \author[$\ast$]{Hyeonju Kim}
-\author[$\ast$]{GregoryFarage}
-\author[$\dagger$]{PjotrPrins}
-\author[$\dagger$]{RobertW.Williams}
-\author[$\S$]{KarlW.Broman}
+\author[$\ast$]{Gregory Farage}
+\author[$\dagger$]{Pjotr Prins}
+\author[$\dagger$]{Robert W. Williams}
+\author[$\S$]{Karl W. Broman}
 \author[$\ast$]{Saunak Sen}
@@
 \subsubsection{Multithreaded CPU operations}
- Used for multi-threaded loops whenever possible, eg. for element-wise operations
+% Used for multi-threaded loops whenever possible, eg. for element-wise operations.
+
 \subsubsection{Matrix and Vectorized Operations}
+
 % OpenBLAS for multi-threaded matrix operations. 
-Matrix multiplication is a well studied area where the computer scientists try to apply various optimization techniques to make it run faster. 
-There are various BLAS libraries available, such as gslBLAS, OpenBLAS, etc. 
-The matrix multiplication in OpenBLAS is default multithreaded, therefore does not require extra coding effort to achieve highly efficient result.
+Since our algorithm largely depends on matrix operations, it is natual to find the fastest way to achieve best result. 
+There are various matrix libraries available, such as gslBLAS and OpenBLAS. 
+They target different hardware or use various techniques to get highly optimal results. 
+The matrix multiplication in OpenBLAS is default multithreaded, therefore does not require extra coding effort to make the CPU version of matrix multiplication run in parallel.
+
+Since matrix multiplication and element-wise operation are algorithmically free of data and function dependency, it is amendable to GPU's parallel computing power. 
+We run matrix multiplication with different shapes matrices, and compared the running time of CPU and GPU. 
+The result is shown in Figure \ref{GPUCPUShape}
+
+\begin{figure}[!htb]
+	\centering
+	\caption{Variation of GPU vs CPU speedup with matrix shape
+	}
+	\label{GPUCPUShape}
+	\includegraphics[scale = 0.36]{figs/speedup.png}
+\end{figure}
 
@@
 %TODO: Add reference to CUDAnative and CuArrays
-\subsubsection{Heritability grid}
-For LMMs, considering a discrete grid of heritabilities enhances parallelization
 
@@
 	\begin{tableminipage}{\textwidth}
-		\begin{tabularx}{\textwidth}{XXXX}
+		\begin{tabularx}{\textwidth}{XXXXX}
 			\hline
-			\header Method & Precision & CPU only   & CPU \& GPU  \\
+			\header Dataset & Method & Precision & CPU only   & CPU \& GPU  \\
 			\hline
-			\multirow{2}{*}{No Find Max LOD} & Single    & 0.36s & 0.41s \\
-			& Double    & 0.55s & 0.80s \\
-			\multirow{2}{*}{Find Max LOD}    & Single    &1.54s       &0.056s       \\
-			& Double    & 1.77s & 0.079s \\
+			\multirow{4}{*}{Spleen Data}                                                & \multirow{2}{*}{No Find Max} & Single    & 0.36s  & 0.41s \\
+			&                              & Double    & 0.55s  & 0.80s \\
+			& \multirow{2}{*}{Find Max}    & Single    & 0.46s  & 0.06s \\
+			&                              & Double    & 0.71s  & 0.08s \\
+			\hline
+			\multirow{2}{*}{\begin{tabular}[c]{@{}c@{}}Hippocampus\\ Data\end{tabular}} & \multirow{2}{*}{Find Max}    & Single    & 17.64s & 2.26s \\
+			&                              & Double    & 26.77s & 3.51s \\
 			\hline
@@
 Our primary concern is the time taken by each function, therefore only timing information is considered in our profiling. 
-We used Julia's built-in Sampling Profiler (%TODO reference here.)
+We used Julia's built-in Sampling Profiler %TODO reference here.)
 to find our target function for GPU because it is less intrusive compared to other profiling methods.
@@
 Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
-Compared to CPU only, which took 1.77 seconds, this method gains 22 times speedup.
-
-\begin{figure}[!htb]
-	\centering
-	\caption{Variation of GPU vs CPU speedup with matrix shape
-	}
-	\label{GPUCPUShape}
-	\includegraphics[scale = 0.36]{figs/speedup.png}
-\end{figure}
+Compared to CPU only, which took 1.77 seconds, this method provides 22 times speedup.
 

### 2019-07-19 paper.tex
wrote matrix mult part in methods

@@
 \usepackage{multirow}
+%\usepackage{listings}
+\newcommand{\code}{\texttt}
 
@@
 % Used for multi-threaded loops whenever possible, eg. for element-wise operations. 
+ Our goal is to build a backend for GeneNetwork that allows researcher to interact with data in real time. 
+ Such requirement sets up the standard that the genome scan must finish within seconds.
+ To bring out the best performance of CPU, we use multithreaded operations whenever possible. 
+ Julia, our choice of programming language, provide simple yet safe syntax for multithreading. 
+ It is simply done by adding \code{Threads.@thread} macro to indicate to Julia that the following \code{for} loop is multhreaded region. 
+ Using \code{Threads.nthreads()} function can show the number of threads in julia, and the default number of threads we use is 16 threads.
 
@@
 % OpenBLAS for multi-threaded matrix operations. 
-Since our algorithm largely depends on matrix operations, it is natual to find the fastest way to achieve best result. 
-There are various matrix libraries available, such as gslBLAS and OpenBLAS. 
+Since our algorithm largely depends on matrix operations, it is natual to find the fastest way to achieve best result, regardless of computing platform. 
+There are various matrix libraries available for CPU, such as gslBLAS and OpenBLAS. 
 They target different hardware or use various techniques to get highly optimal results. 
 The matrix multiplication in OpenBLAS is default multithreaded, therefore does not require extra coding effort to make the CPU version of matrix multiplication run in parallel.
+We chose OpenBLAS as CPU computing library.
 
-Since matrix multiplication and element-wise operation are algorithmically free of data and function dependency, it is amendable to GPU's parallel computing power. 
+Matrix multiplication and element-wise operation are algorithmically free of data and function dependency, it is amendable to GPU's parallel computing power.
+Julia provides various packages for GPU including CUDA, OpenCL bindings. 
+Our chosen hardware for GPU is from Nvidia, which requires their propriatory library CUDA. 
+Besides the hardware requirement, CUDA is also mature and has been recongnized in the scientific computing community. 
+\textit{cuBLAS} library is a fast GPU implementation of the BLAS from Nvidia. 
+For matrix operations on GPU, we used cuBLAS library.
+
+Not only we want to see how much speed up we can get from using GPU, but also experiemnt whether the shape of matrix will affect the speedup. 
+Of course, most of the time, one can not pick and choose the data size or shape he gets, 
+but such information would help researchers as a rough guidance of whether it is worth it to consider the GPU option before investing programming effort for GPU. 
 We run matrix multiplication with different shapes matrices, and compared the running time of CPU and GPU. 
-The result is shown in Figure \ref{GPUCPUShape}
+The experiment setup is, to multiply two input matrices, A(mXn), and B(nXp), and produce an output matrix C(mXp)
+The range of m, n and p is between 2\^4 and 2\^17 in log scale.
+
+The result is shown in Figure \ref{GPUCPUShape}.
 
@@
 However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative and CuArrays affords much lower programming efforts compared to C/C++.
+Compared to writting GPU functions in C/C++, writing in Julia is cleaner and easier, because it requires much less boiler plate code.
 %TODO: Add reference to CUDAnative and CuArrays

### 2019-07-22 paper.tex
update the method

--- 2019-07-18 paper.tex
@@
 	y_i &=& X_j \beta_j+ \epsilon_i,
-	\quad \epsilon_i \sim N(0,\sigma_i^2)
+	\quad \epsilon_i \sim N(0,\sigma_i^2I),
 \end{eqnarray*}
-
 where ${X}_j$ is a matrix that may include the $y-$intercept with
@@
 ($j=1,\ldots,p$).  ${\beta}_j$ is a vector of the $j$-th eQTL
-effects, and ${\epsilon}_i$ is random error.  Suppose $RSS_{0i}$
-is a residual sum of squares under the null hypothesis of no eQTL
-and $RSS_{1ij}$ is the residual sum of squares under the
-alternative for the $i$-th trait and $j$-th marker.  Then the LOD
+effects, and ${\epsilon}_i$ is random error distributed as $N(0,\sigma_i^2I)$.  Suppose $RSS_{0i}$
+is a residual sum of squares under the null hypothesis of no eQTL, and $RSS_{1ij}$ is the residual sum of squares under the
+alternative of existing eQTL at the $i$-th trait and $j$-th marker.  Then the LOD
 score for the corresponding pair can be written as:
@@
 where $r_{ij}$ is the correlation between the $i$-th trait and
-$j$-th marker. Note that this only works for one-df tests.  If the
-trait matrix and genotype matrix ($G$) are standardized, then the
-correlation matrix is simply $$R=Y^{'}G.$$
+$j$-th marker. Note that this only works for one-df tests.  If $Y^{\ast}$ and $G^{\ast}$ are respectively standardized 
+trait ($Y$) and genotype ($G$) matrices, then the
+correlation matrix is simply $$R=Y ^{\ast '}G^{\ast}.$$
 %Indicate which statistical analysis has been performed and describe the method and model applied.

### 2019-07-22 paper.tex
finished writing matrix multiplication experiment, and merge in hyeonju edits

@@
 Not only we want to see how much speed up we can get from using GPU, but also experiemnt whether the shape of matrix will affect the speedup. 
-Of course, most of the time, one can not pick and choose the data size or shape he gets, 
+Of course, most of the time, one can not pick and choose the size and shape of data in matrix form, 
 but such information would help researchers as a rough guidance of whether it is worth it to consider the GPU option before investing programming effort for GPU. 
 We run matrix multiplication with different shapes matrices, and compared the running time of CPU and GPU. 
+CPU time is matrix multiplication from OpenBLAS library using 16 threads. 
+GPU time includes all overhead of using GPU, which involves device launch, data transfer and all necessary API calls.
+We think to make a fair comparison between CPU and GPU, we need to use maximum strength of both, and include all necessary cost.
+
 The experiment setup is, to multiply two input matrices, A(mXn), and B(nXp), and produce an output matrix C(mXp)
 The range of m, n and p is between 2\^4 and 2\^17 in log scale.
-
+Due to the GPU hardware limitations, we can only compare the result when the size of input and output (I/O size) matrix total is less than 16 GB. 
 The result is shown in Figure \ref{GPUCPUShape}. 
+The X axis of Figure \ref{GPUCPUShape} is the dimensions of matrix whose I/O size is over 12 GB.
+The Y axis is the speedup of GPU comparing to CPU in log scale.  
+From this figure, the matrices whose shape are closer to a square matrix, get better speedup from GPU.
 

### 2019-07-22 paper.tex
adding julia gpu code sample

@@
 %\usepackage{listings}
+\usepackage{minted}
 \newcommand{\code}{\texttt}
@@
 The experiment setup is, to multiply two input matrices, A(mXn), and B(nXp), and produce an output matrix C(mXp)
-The range of m, n and p is between 2\^4 and 2\^17 in log scale.
+The range of m, n and p is between 2\textsuperscript{4} and 2\textsuperscript{17} in log scale.
 Due to the GPU hardware limitations, we can only compare the result when the size of input and output (I/O size) matrix total is less than 16 GB. 
@@
 The X axis of Figure \ref{GPUCPUShape} is the dimensions of matrix whose I/O size is over 12 GB.
+This is because using GPU involves a lot of overhead. 
+Such overhead can only be justified by processing large data sets. 
 The Y axis is the speedup of GPU comparing to CPU in log scale.  
@@
 Compared to writting GPU functions in C/C++, writing in Julia is cleaner and easier, because it requires much less boiler plate code.
+
+\inputminted{julia}{code/gpu.jl}
 %TODO: Add reference to CUDAnative and CuArrays
@@
 Marker data was available on 7321 markers; genotype probabilities were calculated using R/qtl.
+%TODO: add hippocampus data description.
 

### 2019-07-22 code/gpu.jl

@@ -1,0 +1,18 @@
+# Calling GPU BLAS library
+CuArrays.CUBLAS.gemm('T', 'N', a,b);
+
+# Writing custom kernel for calculating LOD score 
+# from correlation matrix
+function lod_kernel(input, MAX,n)
+  i = (blockIdx().x-1) * blockDim().x + threadIdx().x
+  if(i < MAX+1)
+    r_square = (input[i]/n)^2
+    input[i] = (-n/Float64(2.0)) *  
+               CUDAnative.log(Float64(1.0)-r_square)
+  end
+  return
+end
+
+# Launching GPU
+@cuda blocks=1000 threads=64 
+lod_kernel(input, MAX, n)

### 2019-07-23 paper.tex
added software availability section

@@
 From this figure, the matrices whose shape are closer to a square matrix, get better speedup from GPU.
+Matrix multiplication is up to 5 times faster on GPU compared to 16 threaded CPU.
 
@@
 \subsubsection{GPU operations}
-\iffalse
-% this block is commented out 
-	In order to achieve the goal of real time eQTL scans, we profiled the entire process of eQTL scan.
-Profiling can provide insightful information about the program,  such as running time of each function.
-	Targeting the longest running time functions provides the most gain with limited programming effort and computing resources. 
-	In our profiling result, we found that the is calculating correlation matrix and LOD score takes up to 90\% of total running time. 
-	Since both steps uses matrix multiplication, which is highly parallelizable and well-researched, we choose the following approaches to speed up matrix multiplication. 
-	First, we chose OpenBLAS library for matrix multiplication, which is an optimized  multithreaded BLAS library. 
-	In our experiment, gslBLAS (GNU Scientific Library) is 10 times slower than OpenBLAS. 
-	In addition to choosing an optimized multithreaded library, we use multithreading operations whenver possible.
-	Second, we considered the possibility of using single precision compared to double precision. 
-	Precision change will affect both throughput and data transfer time. 
-	Throughput is the number of operations done in unit time, usually measured in FLOPS(floating point operations).
-	In our experiment, single precision is twice as fast as double precision, and provides correct result within 1e-2 tolerance. 
-	Detailed running time of single precision and double precision is show in Table 1. 
-	Third, we considered GPU as an attractive computing platform, because of its massive number of cores at a lower price range, and fast open source GPGPU libraries such as CUDA and OpenCL. 
-	Matrix multiplication and element-wise operations are amendable to GPU's heterogeneous computing architecture, since both have no data race conditions and low data dependencies. 
-	In our experiment, matrix multiplication is up to 5 times faster compared to 16 threaded CPU. 
-	We compared the speed of matrix multiplication on CPU and GPU, with rectangular matrix sizes ranging from 1k to 100k, increased by power of 2, 
-	We found speed up is sensitive to the size of matrix, and the shape of matrix.  
-	GPU is best used for large number of operations, therefore multiplying larger matrices will give better speed up. 
-	The shape of matrices also plays an important role when the matrix size is large enough. 
-	Figure 2 shows the speed up of matrix multiplication, when the total data size of input and output floating point matrices is over 12 GB. 
-	Our reasoning for such sensitivity is that the data transfer from GPU(device) to CPU(host) is half of CPU to GPU. 
-	When output data size is larger than input matrix, the speed up will decrease. 
-	Further experiment will be done to prove such hypothesis. 
-	In addition, our GPU profiling result shows that 98\% of eQTL scan for BXD spleen dataset is spent on data transfer for the GPU approach. 
-	Minimizing data transfer time will make the scan significantly faster. 
-	Further programming efforts will make GPU a more attractive approach to reach real time eQTL scans. 
-	Lastly but most fundamentally, we choose Julia programming language as our development tool. 
-	In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
-	However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative and CuArrays affords much lower programming efforts compared to C/C++.
-\fi
 
@@
 Compared to writting GPU functions in C/C++, writing in Julia is cleaner and easier, because it requires much less boiler plate code.
+Below is some code snippets to demonstrate GPU code, calling library or writing custom kernel in Julia.
 
@@
 \subsection{Information and Availability of Datasets}
-Data from 79 BXD strains for 35556 transcripts from spleens were collected using the Affy Mouse Gene 1.0 ST array.
+
+We used 2 BXD data sets of mouse, spleen and hippocampus data sets to test our method. 
+Data from 79 BXD strains for 35556 transcripts from spleens were collected using the Affy Mouse Gene 1.0 ST array. GN accession: GN283. Marker data was available on 7321 markers;
+Hippocampus data for 93 strains (70 BXDs) summarized at the exon level for 1,236,087 probes/probe sets.
+GeneNetwork Accession Number: GN206
 Animals were generated by Drs.
@@
 All animals were young adults (oldest 160 days).
-GN accession: GN283.
-Marker data was available on 7321 markers; genotype probabilities were calculated using R/qtl.
-%TODO: add hippocampus data description.
+genotype probabilities were calculated using R/qtl.
 
@@
 \section{Results}
+
 \subsection{Benefit of using linear model}
@@
 , while R matrix is about 4GB.
-For our test hardware Nvidia TESLA V100, data transfer rate from CPU to GPU (host to device) is 4.3GB/s, and GPU to CPU (device to host) is 2.2GB/s. %TODO: double check the numbers here. 
-Larger data size at a slower transfer rate makes GPU seems unattractive for computing genome scan, at least for BXD dataset.
+Data offloading is the main bottleneck for our GPU implementation.
 
@@
 This step is highly parallelizable and can utilize GPU's massive cores, at the same time, reduce the amount of data needed to be transfered back to host. 
-Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
-Compared to CPU only, which took 1.77 seconds, this method provides 22 times speedup.
+Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method for spleen data reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
+Compared to CPU only, which took 0.55 seconds, this GPU modification provides 7 times speedup. 
+For hippocampus datasets, which contains over 1 million phenotypes, the GPU Find Max implementation took 3.51 seconds for double precision, and 2.26 seconds for single precision. 
+As a backend service to a website, this wait time is much more preferable than the CPU timing for real time interactions.
 
-\iffalse
-We profiled the entire genome scan program to get a complete picture of the execution process.
-Our profiling result shows that 90\% of runtime is spent on calculating correlation matrix and LOD score. 
-Since both involve matrix operations (multiplication and elementwise operations respectively) which are highly parallelizable, we used GPU to speed up matrix operations.
+\subsection{Software Availability}
+The source code for this package is publically available on Bitbucket: https://bitbucket.org/Xiaoqi-hu/lm-gpu.
+This repository contains three folders: \textit{data}, \textit{src} and \textit{test}. 
+Using Julia language and running \code{test.jl} in \textit{test} folder will run our method using either spleen data and hippocampus data in \textit{data} folder.
 
-{\em Profiling:} Calculating correlation matrix and LOD score
-takes up to 90\% of runtime.  Both involve matrix operations
-(multiplication and elementwise operations respectively) which
-are highly parallelizable.
-      \begin{itemize}
-	\item We chose OpenBLAS library for matrix multiplication, an
-	optimized multithreaded library.  In our experiment, gslBLAS
-	(GNU Scientific Library) is 10 times slower than OpenBLAS.
-	\item We used multithreading operations whenever possible, such as
-	for elementwise operations.
-	\item We explored using single precision instead of double
-	precision.  Precision change affects both throughput (FLOPS) and
-	data transfer time.  Single precision is twice
-	as fast as double precision, and provides correct result	 
-	within 1e-2 tolerance (Table 1).
-\end{itemize}
-
- {\em GPUs vs CPUs:} GPUs are attractive since they provide a
-massive number of cores at a lower price range.
-Matrix multiplication and element-wise operations are amendable
-to GPU's heterogeneous computing architecture, since both have
-no data race conditions and low data dependencies.
-
-\begin{itemize}
-	\item Matrix multiplication is up to 5 times faster
-	compared to 16-threaded CPU.
-	% We compared the speed of matrix
-	% multiplication on CPU and GPU, with matrix sizes
-	% ranging from 1k to 100k.
-	We found speed
-	up is sensitive to the size and the shape of matrices.
-	% GPU is best used for large number of operations, therefore
-	% multiplying larger matrices will give better speed up.  The
-	% shape of matrices also plays an important role when the matrix
-	% size is large enough.
-	Figure 2 shows the speedup of matrix
-	multiplication, when the total data size of input and output
-	floating point matrices is over 12 GB.
-	% Our reasoning for such
-	% sensitivity is that the data transfer from GPU(device) to
-	% CPU(host) is half of CPU to GPU.
-	When output data size is
-	larger than input matrix, the speed up will decrease. 
-	\item  GPU profiling shows that 98\% of eQTL scan for BXD
-	spleen dataset is spent on data transfer.
-	Minimizing data transfer time will make the scan significantly
-	faster.
-\end{itemize}
-
-      {\em Julia as development platform:} In our experiment
-of matrix multiplication, Julia's speed is comparable to C/C++.
-However, the low learning curve, clean syntax, as well as
-support to GPU programming libraries such as CUDAnative and
-CuArrays affords much lower programming efforts compared to
-C/C++.
-
-\fi
+This package is developed in Julia language, uses openBLAS and CUDA10.1. 
+To use this package, one needs to have an Nvidia GPU in the test machine,
+and have Julia language binary, openBLAS and CUDA library installed.
 
 \subsection{Platform}
+Hardware:
  \begin{itemize}
@@
 
-%The results and discussion should not be repetitive.
-The results section should give a factual presentation of the data and all tables and figures should be referenced; the discussion should not summarize the results but provide an interpretation of the results, and should clearly delineate between the findings of the particular study and the possible impact of those findings in a larger context.
-Authors are encouraged to cite recent work relevant to their interpretations.
-Present and discuss results only once, not in both the Results and Discussion sections.
-It is sometimes acceptable to combine results and discussion.
-The text should be as succinct as possible.
-Heed Strunk and White's dictum: "Omit needless words!"
-
 \section{Conclusion}

### 2019-07-24 paper.tex
minor modification

@@
 
-\title{Real Time Genome Scan Using GPU}
+\title{Speeding up eQTL scans in the BxD population using GPU}
 
@@
 \author[$\S$]{Karl W. Broman}
-\author[$\ast$]{Saunak Sen}
+\author[$\ast$,1]{Saunak Sen}
 
@@
 
-\runningtitle{G3 Journal Template on Overleaf} % For use in the footer
+\runningtitle{Speeding up eQTL scans in the BxD population using GPU} % For use in the footer
 
@@
 % contributions, using the relevant superscript numbers
-\equalcontrib{1}
-\equalcontrib{2}
+%\equalcontrib{1}
+%\equalcontrib{2}
 
-\correspondingauthoraffiliation{3}{Corresponding author: Please insert the affiliation correspondence address and email for the corresponding author.
-The corresponding author should be marked with the relevant number in the author list, as shown in the example.}
+\correspondingauthoraffiliation{1}{Corresponding author: Saunak Sen, Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis, TN. sen@uthsc.edu}
 \vspace{-34pt}% Only used for adjusting extra space in the left column of the first page
@@
 populations such as the BXD population.
+
+Compare matrix eQTL \citep{shabalin2012matrix}, GridLMM \citep{runcie2018fast}, Bolt-LMM \citep{loh2015efficient}
 
@@
  To bring out the best performance of CPU, we use multithreaded operations whenever possible. 
- Julia, our choice of programming language, provide simple yet safe syntax for multithreading. 
+ Julia \citep{bezanson2017julia}, our choice of programming language, provide simple yet safe syntax for multithreading. 
  It is simply done by adding \code{Threads.@thread} macro to indicate to Julia that the following \code{for} loop is multhreaded region. 
@@
 Since our algorithm largely depends on matrix operations, it is natual to find the fastest way to achieve best result, regardless of computing platform. 
-There are various matrix libraries available for CPU, such as gslBLAS and OpenBLAS. 
+There are various matrix libraries available for CPU, such as gslBLAS and OpenBLAS \citep{wang2013augem}. 
 They target different hardware or use various techniques to get highly optimal results. 
@@
 Matrix multiplication and element-wise operation are algorithmically free of data and function dependency, it is amendable to GPU's parallel computing power.
-Julia provides various packages for GPU including CUDA, OpenCL bindings. 
+Julia provides various packages for GPU including CUDA \citep{Nickolls:2008:SPP:1365490.1365500}, OpenCL bindings. 
 Our chosen hardware for GPU is from Nvidia, which requires their propriatory library CUDA. 
 Besides the hardware requirement, CUDA is also mature and has been recongnized in the scientific computing community. 
-\textit{cuBLAS} library is a fast GPU implementation of the BLAS from Nvidia. 
+\textit{cuBLAS} \citep{cublas} library is a fast GPU implementation of the BLAS from Nvidia. 
 For matrix operations on GPU, we used cuBLAS library.
@@
 	\label{GPUCPUShape}
-	\includegraphics[scale = 0.36]{figs/speedup.png}
+	\includegraphics[scale = 0.35]{figs/speedup.png}
 \end{figure}
@@
 In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
-However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative and CuArrays affords much lower programming efforts compared to C/C++.
+However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative \citep{CUDAnative.jl-2018} and CuArrays affords much lower programming efforts compared to C/C++.
 Compared to writting GPU functions in C/C++, writing in Julia is cleaner and easier, because it requires much less boiler plate code.
@@
 Our primary concern is the time taken by each function, therefore only timing information is considered in our profiling. 
-We used Julia's built-in Sampling Profiler %TODO reference here.)
-to find our target function for GPU because it is less intrusive compared to other profiling methods.
+We used Julia's built-in Sampling Profiler to find our target function for GPU because it is less intrusive compared to other profiling methods.
 
@@
 In contrast, it is slower, whether using single or double precision for the data type.
-We used a GPU profiler \textit{nvprof} %TODO reference here
+We used a GPU profiler \textit{nvprof} \cite{nvprof}
 to investigate what is the bottleneck of GPU. 
@@
 \subsection{Software Availability}
-The source code for this package is publically available on Bitbucket: https://bitbucket.org/Xiaoqi-hu/lm-gpu.
+The source code for this package is publically available on Bitbucket: \url{https://bitbucket.org/Xiaoqi-hu/lm-gpu}.
 This repository contains three folders: \textit{data}, \textit{src} and \textit{test}. 
@@
 	\item Libraries: CUDA v10.1 and cuBLAS; OpenBLAS
-	\item Profilers: Julia Profiler Package; nvprof
+	\item Profilers: Julia Profiler; nvprof
 \end{itemize}
@@
 
-\bibliography{example-bibliography}
+LM-GPU provides a fast backend for GeneNetwork that is powered by GPU. 
+It enables researchers to do real-time interaction with the data provided in GeneNetwork. 
+Such possibility can help researchers experiment with different parameters, and results in insightful findings. 
+LM-GPU takes advantage of GPU's massive parallel computing power and Julia's elegant syntax for quick prototyping and implementing. 
+We are able to run genome scan for hippocampus data with over one million phenotypes in three seconds.
+
+%TODO: limitations. One degree of freedom
+
+%Further development of LM-GPU is
+
+\bibliography{reference}
 

### 2019-07-25 paper.tex
Added to the intro.

@@
 \author[$\S$]{Karl W. Broman}
-\author[$\ast$,1]{Saunak Sen}
+\author[$\ast$,1]{\'Saunak Sen}
 
@@
 
-\runningtitle{Speeding up eQTL scans in the BxD population using GPU} % For use in the footer
+\runningtitle{Speeding up eQTL scans in the BXD population using GPU} % For use in the footer
 
@@
 Julia programming language.
-
 \end{abstract}
@@
 
-\noindent        Computational demands in omics are growing exponentially. For
-instance, eQTL analysis requires processing genomewide
-expression and marker data; this can take minutes, hours or
-days depending on problem size.  Scientists want to intereact
-with the results and real-time processing would be valuable.
-Our goal is to build a backend for GeneNetwork that will enable
-real-time genome scans for important genetic reference
-populations such as the BXD population.
+\noindent The BXD mouse strains are an important set of recombinant
+inbred lines with deep phenotypic characterization.  Since the strains
+have been genotyped and sequenced, once a set of strains is
+phenotyped, the data can be immediately used for QTL (quantitative
+trait locus) mapping, and associational analysis with
+previously-collected phenotypes.  For ``omic'' phenotypes collected
+using high-throughput technologies additional analyses (such as
+transcriptional network construction or causal mediation analyses) are
+possible.  The GeneNetwork project ({\tt http://gn2.genenetwork.org})
+seeks to faculitate this process by providing a searchable database of
+phenotypes and genotypes in a variety of organisms (including mouse,
+rat, and Arabidopsis).  It provides a suite of interactive tools for
+browsing the data, QTL mapping, correlational analyses, network
+construction, and genome browsing.  Our goal is to develop a real-time
+backend for GeneNetwork to perform eQTL analysis (where we perform
+single-trait QTL mapping for thousands of ``omic'' traits) for
+important genetic reference populations such as the BXD population.
+
+%% Computational demands in omics are growing exponentially. For
+%% instance, eQTL analysis requires processing genomewide expression and
+%% marker data; this can take minutes, hours or days depending on problem
+%% size.  Scientists want to intereact with the results and real-time
+%% processing would be valuable.  Our goal is to build a backend for
+%% GeneNetwork that will enable real-time genome scans for important
+%% genetic reference populations such as the BXD population.
 

### 2019-07-25 paper.tex
BXD not BxD.
Added bibtex command to Makefile.
made small changes to manuscript

@@
 
-\title{Speeding up eQTL scans in the BxD population using GPU}
+\title{Speeding up eQTL scans in the BXD population using GPU}
 
@@
 
-The experiment setup is, to multiply two input matrices, A(mXn), and B(nXp), and produce an output matrix C(mXp)
+The experiment setup is, to multiply two input matrices, A(m$\times$n), and B(n$\times$p), and produce an output matrix C(m$\times$p)
 The range of m, n and p is between 2\textsuperscript{4} and 2\textsuperscript{17} in log scale.
@@
 % Faster and memory efficient; may be less accurate, but acceptable for quick scan
-Precision, in computer science, means the smallest difference between two representable numbers.
+Precision means the smallest difference between two representable numbers.
 Floating point numbers, in scientific computing, are usually stored in double presicision. 
@@
 Differences in optimization techniques, underlying architecture cause such performance gap. 
-Considering the memory size, data transfer speed and throughput, single precision brings multiple benefits when precision is not the priority concern.
+Considering the storage size, data transfer speed and throughput, single precision brings multiple benefits when precision is not the priority concern.
 
@@
 Compared to CPU only, which took 0.55 seconds, this GPU modification provides 7 times speedup. 
+As we mentioned earlier, since we are porting the computation to GPU which took 90\% of total execution time, according to Amdahl's Law, the theoretical maximum speed up is 10 times. 
+We got 7 times speed up for genome scan using GPU. 
 For hippocampus datasets, which contains over 1 million phenotypes, the GPU Find Max implementation took 3.51 seconds for double precision, and 2.26 seconds for single precision. 

### 2019-07-26 paper.tex
Added to intro.

@@
 
-Compare matrix eQTL \citep{shabalin2012matrix}, GridLMM \citep{runcie2018fast}, Bolt-LMM \citep{loh2015efficient}
+To perform eQTL scans in the BXD population, one has to simply perform
+as many genome scans as there are phenotypes.  This can be done in an
+``embarassingly parallel'' fashion by using standard algorithms for
+QTL analysis, such as those employed by R/qtl (cite).  In practice,
+this is too slow, and substantial speedups are possible.  [Mention
+  ideas that have been used before] For example, by using the
+Haley-Knott algorithms using genotype probabilities on batches of
+phenotypes with the same missing data pattern, instead of using the EM
+algorithm on each phenotype individually, substantial speedups are
+possible.  This is a well-known trick and is used by R/qtl.
+Additionally, if only additive effects are tested, or if the
+population has only two genotype categories (as in a backcross or
+recombinant inbred line), then matrix multiplication can be used to
+perform Haley-Knott regression \citep{shabalin2012matrix}.
+
+%% Compare matrix eQTL GridLMM \citep{runcie2018fast}, Bolt-LMM
+%% \citep{loh2015efficient}
+
+We sought to build upon these efforts aiming to perform real-time eQTL
+scans for the BXD populations.  We used the above-mentioned ideas and
+examined the use of graphical processing units (GPUs) to speed these
+further.  Since programming for GPUs is often non-trivial needing the
+use of low-level languages such as C++, we used the Julia programming
+language which offers a GPU programming capabilities while retaining
+the simplicity of a high-level language such as R or Matlab.  Finally,
+since most phenotype-marker associations are null, we examined the
+impact of storage precision, and only returning the highest
+association (LOD) score for each phenotype instead of a matrix of LOD
+scores for every pair of marker and phenotypes.  With these efforts,
+we find that it is possible to perform perform real-time eQTL scans
+for the BXD population.
 
@@
 If working with populations indicate how samples were collected and whether they were random with respect to the target population.
+
+\subsection{BXD genotypes and phenotypes}
+
+Describe genotype and phenotype data here.
 

### 2019-07-30 paper.tex
Rob edited abstract of paper.tex online with Bitbucket

@@
 \affil[$\ast$]{Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis, TN} 
-\affil[$\dagger$]{Department of Genetics, Genomics, and Informatics, University of Tennessee Health Science Center, Memphis, TN} 
+\affil[$\dagger$]{Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, TN} 
 \affil[$\S$]{Department of Biostatistics, University of Wisconsin-Madison, Madison, WI}
@@
 \begin{abstract}
-The BXD strains are an important reference population of
-recombinant inbred lines which have been phenotyped extensively.
+The BXD strains of mice are an important reference population for
+systems biology and genetics that have been full sequenced and deeply phenotyped.
 To facilitate interactive use of genotype-phenotype relationships
-in this population, we sought to speed up eQTL scans where we
-perform a univariate genome scan for every trait in a collection
-of omic traits.  By using easily parallelizable operations such as
+in this family, we sought to develop new algorithms and code to enable
+high performance whole genome QTL scan for massive omics data sets of up to 1 million 
+traits. By using easily parallelizable operations including
 matrix multiplication, vectorized operations, and elementwise
-operations, we are able to decrease runtimes approaching real-time
-computation.  We used parallelization using different CPU threads
+operations, we have been able to decrease runtimes approaching real-time
+computation, making this code ideal for interactive web services, such as
+GeneNetwork.org. We used parallelization of different CPU threads
 as well as GPUs. We found that the speed advantage of GPUs is
-dependent on problem size and shape.  These results indicate a pathway for
-speeding up eQTL scans using LMMs.  Our implementation is in the
-Julia programming language.
+dependent on problem size and shape (n of cases, n of genotypes, n of traits).  
+Our results provide a clear path for speeding up eQTL scans using LMMs. 
+Our implementation is in the Julia programming language.
 \end{abstract}

### 2019-07-31 paper.tex
Typos

@@
 %\usepackage{listings}
-\usepackage{minted}
+\usepackage[cache=false]{minted}
 \newcommand{\code}{\texttt}
@@
 possible.  The GeneNetwork project ({\tt http://gn2.genenetwork.org})
-seeks to faculitate this process by providing a searchable database of
+seeks to facilitate this process by providing a searchable database of
 phenotypes and genotypes in a variety of organisms (including mouse,
@@
 
-Not only we want to see how much speed up we can get from using GPU, but also experiemnt whether the shape of matrix will affect the speedup. 
+Not only we want to see how much speed up we can get from using GPU, but also experiment whether the shape of matrix will affect the speedup. 
 Of course, most of the time, one can not pick and choose the size and shape of data in matrix form, 
@@
 Floating point numbers, in scientific computing, are usually stored in double presicision. 
-Double precision floating point numebrs takes up 8 bytes in memory, while single precision takes up 4 bytes. 
+Double precision floating point numbers takes up 8 bytes in memory, while single precision takes up 4 bytes. 
 In addition to the difference in storing size, the speed for calculation using single and double precision also varies depending on hardware. 
@@
 
-To cope with this limitaion of GPU, we further developed \textit{Find Max LOD} method, because main interest of genome scan is to find the maximum LOD score of every phenotype. 
+To cope with this limitation of GPU, we further developed \textit{Find Max LOD} method, because main interest of genome scan is to find the maximum LOD score of every phenotype. 
 This step is highly parallelizable and can utilize GPU's massive cores, at the same time, reduce the amount of data needed to be transfered back to host. 

### 2019-08-09 paper.tex
Fix some typos

@@
 %% marker data; this can take minutes, hours or days depending on problem
-%% size.  Scientists want to intereact with the results and real-time
+%% size.  Scientists want to interact with the results and real-time
 %% processing would be valuable.  Our goal is to build a backend for
@@
 as many genome scans as there are phenotypes.  This can be done in an
-``embarassingly parallel'' fashion by using standard algorithms for
+``embarrassingly parallel'' fashion by using standard algorithms for
 QTL analysis, such as those employed by R/qtl (cite).  In practice,
@@
 		the expression phenotype matrix, $G$ is the matrix of
-		genotytpes, and $R=Y^{\prime}G$ is the matrix of
+		genotypes, and $R=Y^{\prime}G$ is the matrix of
 		correlations.  The LOD scores are a function of the
@@
  Julia \citep{bezanson2017julia}, our choice of programming language, provide simple yet safe syntax for multithreading. 
- It is simply done by adding \code{Threads.@thread} macro to indicate to Julia that the following \code{for} loop is multhreaded region. 
+ It is simply done by adding \code{Threads.@thread} macro to indicate to Julia that the following \code{for} loop is multithreaded region. 
  Using \code{Threads.nthreads()} function can show the number of threads in julia, and the default number of threads we use is 16 threads.
@@
 % OpenBLAS for multi-threaded matrix operations. 
-Since our algorithm largely depends on matrix operations, it is natual to find the fastest way to achieve best result, regardless of computing platform. 
+Since our algorithm largely depends on matrix operations, it is natural to find the fastest way to achieve best result, regardless of computing platform. 
 There are various matrix libraries available for CPU, such as gslBLAS and OpenBLAS \citep{wang2013augem}. 
@@
 
-Matrix multiplication and element-wise operation are algorithmically free of data and function dependency, it is amendable to GPU's parallel computing power.
+Matrix multiplication and element-wise operation are algorithmically free of data and function dependency, it is amenable to GPU's parallel computing power.
 Julia provides various packages for GPU including CUDA \citep{Nickolls:2008:SPP:1365490.1365500}, OpenCL bindings. 
-Our chosen hardware for GPU is from Nvidia, which requires their propriatory library CUDA. 
-Besides the hardware requirement, CUDA is also mature and has been recongnized in the scientific computing community. 
+Our chosen hardware for GPU is from Nvidia, which requires their proprietary library CUDA. 
+Besides the hardware requirement, CUDA is also mature and has been recognized in the scientific computing community. 
 \textit{cuBLAS} \citep{cublas} library is a fast GPU implementation of the BLAS from Nvidia. 
@@
 Precision means the smallest difference between two representable numbers.
-Floating point numbers, in scientific computing, are usually stored in double presicision. 
+Floating point numbers, in scientific computing, are usually stored in double precision. 
 Double precision floating point numbers takes up 8 bytes in memory, while single precision takes up 4 bytes. 
@@
 Based on our profiling result, the time consuming parts of our genome scan method are matrix multiplication and element-wise operations. 
-Both are amendable to GPU's heterogeneous computing architecture, since they have no data race conditions and low data dependencies. 
+Both are amenable to GPU's heterogeneous computing architecture, since they have no data race conditions and low data dependencies. 
 However, GPU also has its own limitations. 
-To truely utilize the maximum computing power of GPU, one needs to think creatively to work around such limitations from GPU. 
-For example, during our experiemnt, memory transfer between host and device is really slow. 
+To truly utilize the maximum computing power of GPU, one needs to think creatively to work around such limitations from GPU. 
+For example, during our experiment, memory transfer between host and device is really slow. 
 Profiling result shows that 98\% of total genome scan time is spent on memory transfer. 
@@
 However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative \citep{CUDAnative.jl-2018} and CuArrays affords much lower programming efforts compared to C/C++.
-Compared to writting GPU functions in C/C++, writing in Julia is cleaner and easier, because it requires much less boiler plate code.
+Compared to writing GPU functions in C/C++, writing in Julia is cleaner and easier, because it requires much less boiler plate code.
 Below is some code snippets to demonstrate GPU code, calling library or writing custom kernel in Julia.
@@
 For example, if a program took ten minutes for a serial processor, and a function that takes nine our of those ten minutes can be parallelized, then, the theoretical speedup, no matter how many processors are used, can not be more than ten times, because the minimum execution time of this program is one minute.
-Therefore, profiling the entire genome scan process is the prerequisit in order to begin the optimization. 
+Therefore, profiling the entire genome scan process is the prerequisite in order to begin the optimization. 
 Often, profiling would consider space and time complexity. 
@@
 \begin{itemize}
-	\item Calculate standadized matrix for input matrices.
+	\item Calculate standardized matrix for input matrices.
 	\item Get correlation matrix
-	\item Caclulate LOD score
+	\item Calculate LOD score
 \end{itemize}
 Our profiling result shows that step two and three takes up 90\% of time, and both involve parallelizable matrix operations. 
-So they are our candidates for GPU accleration.
+So they are our candidates for GPU acceleration.
 To make fair comparison for CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
@@
 To cope with this limitation of GPU, we further developed \textit{Find Max LOD} method, because main interest of genome scan is to find the maximum LOD score of every phenotype. 
-This step is highly parallelizable and can utilize GPU's massive cores, at the same time, reduce the amount of data needed to be transfered back to host. 
+This step is highly parallelizable and can utilize GPU's massive cores, at the same time, reduce the amount of data needed to be transferred back to host. 
 Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method for spleen data reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
@@
 \subsection{Software Availability}
-The source code for this package is publically available on Bitbucket: \url{https://bitbucket.org/Xiaoqi-hu/lm-gpu}.
+The source code for this package is publicly available on Bitbucket: \url{https://bitbucket.org/Xiaoqi-hu/lm-gpu}.
 This repository contains three folders: \textit{data}, \textit{src} and \textit{test}. 

### 2019-08-10 paper.tex
corrected typos and added data info

--- 2019-07-31 paper.tex
@@
 previously-collected phenotypes.  For ``omic'' phenotypes collected
-using high-throughput technologies additional analyses (such as
+using high-throughput technologies, additional analyses (such as
 transcriptional network construction or causal mediation analyses) are
@@
 ``embarassingly parallel'' fashion by using standard algorithms for
-QTL analysis, such as those employed by R/qtl (cite).  In practice,
+QTL analysis such as those employed by R/qtl (cite).  In practice,
 this is too slow, and substantial speedups are possible.  [Mention
@@
 
-\subsection{BXD genotypes and phenotypes}
+% This data subsection is repeated in the last subsection (information and availability of Data sets)
+%\subsection{BXD genotypes and phenotypes}
 
-Describe genotype and phenotype data here.
+%Describe genotype and phenotype data here.
 
@@
 \subsection{Acceleration Techniques}
-We used various techniques to accelerate eQTL scans. 
+We used following techniques to accelerate eQTL scans. 
 \subsubsection{Multithreaded CPU operations}
@@
 They target different hardware or use various techniques to get highly optimal results. 
-The matrix multiplication in OpenBLAS is default multithreaded, therefore does not require extra coding effort to make the CPU version of matrix multiplication run in parallel.
+The matrix multiplication in OpenBLAS is default multithreaded and therefore does not require extra coding effort to make the CPU version of matrix multiplication run in parallel.
 We chose OpenBLAS as CPU computing library.
 
-Matrix multiplication and element-wise operation are algorithmically free of data and function dependency, it is amendable to GPU's parallel computing power.
+Matrix multiplication and element-wise operation are algorithmically free of data and function dependency, so that they are amendable to GPU's parallel computing power.
 Julia provides various packages for GPU including CUDA \citep{Nickolls:2008:SPP:1365490.1365500}, OpenCL bindings. 
-Our chosen hardware for GPU is from Nvidia, which requires their propriatory library CUDA. 
+Our chosen hardware for GPU is from Nvidia, which requires its propriatory library CUDA. 
 Besides the hardware requirement, CUDA is also mature and has been recongnized in the scientific computing community. 
@@
 
-Not only we want to see how much speed up we can get from using GPU, but also experiment whether the shape of matrix will affect the speedup. 
-Of course, most of the time, one can not pick and choose the size and shape of data in matrix form, 
-but such information would help researchers as a rough guidance of whether it is worth it to consider the GPU option before investing programming effort for GPU. 
-We run matrix multiplication with different shapes matrices, and compared the running time of CPU and GPU. 
-CPU time is matrix multiplication from OpenBLAS library using 16 threads. 
+Not only do we want to see how much speed up we can get from using GPU, but we also experiment whether the shape of matrix will affect the speedup. 
+Of course, most of the time, one can not pick the size and shape of data in a matrix form, 
+but such information would help researchers as a rough guidance of whether it is worth considering the GPU option before investing programming effort for GPU. 
+We run matrix multiplication with different shapes of matrices, and compared the running time of CPU and GPU. 
+CPU time is measured in matrix multiplication from OpenBLAS library using 16 threads. 
 GPU time includes all overhead of using GPU, which involves device launch, data transfer and all necessary API calls.
-We think to make a fair comparison between CPU and GPU, we need to use maximum strength of both, and include all necessary cost.
+In order to make a fair comparison between CPU and GPU, we need to use maximum strength of both and include all necessary cost.
 
-The experiment setup is, to multiply two input matrices, A(m$\times$n), and B(n$\times$p), and produce an output matrix C(m$\times$p)
+The experiment setup is to multiply two input matrices, A(m$\times$n), and B(n$\times$p), and produce an output matrix C(m$\times$p).  
 The range of m, n and p is between 2\textsuperscript{4} and 2\textsuperscript{17} in log scale.
-Due to the GPU hardware limitations, we can only compare the result when the size of input and output (I/O size) matrix total is less than 16 GB. 
+Due to the GPU hardware limitations, we can only compare the result when the size of input and output (I/O size) matrix, in total, is less than 16 GB. 
 The result is shown in Figure \ref{GPUCPUShape}. 
@@
 Such overhead can only be justified by processing large data sets. 
-The Y axis is the speedup of GPU comparing to CPU in log scale.  
-From this figure, the matrices whose shape are closer to a square matrix, get better speedup from GPU.
-Matrix multiplication is up to 5 times faster on GPU compared to 16 threaded CPU.
+The Y axis is the speedup of GPU compared to CPU in log scale.  
+From this figure, the matrices whose shapes are closer to square matrices get better speedups from GPU.
+Matrix multiplication is up to 5 times faster on GPU compared to that on 16 threaded CPU.
 
@@
 Floating point numbers, in scientific computing, are usually stored in double presicision. 
-Double precision floating point numbers takes up 8 bytes in memory, while single precision takes up 4 bytes. 
-In addition to the difference in storing size, the speed for calculation using single and double precision also varies depending on hardware. 
-For example, the GPU throughput (number of floating point calculation per second, measured in FLOPS) for double precision is 1/32 of single precision on a Nvidia GTX 1050 GPU, and 1/4 on a Nvidia Tesla K80.
+Double precision takes up 8 bytes in memory, while single precision takes up 4 bytes. 
+In addition to the difference in storing size, the speed for calculation using single and double precisions also varies 
+depending on the hardware.  For example, the GPU throughput (number of floating point calculation per second, measured in FLOPS)
+for double precision is 1/32 of single precision on a Nvidia GTX 1050 GPU, and 1/4 on a Nvidia Tesla K80. 
 Differences in optimization techniques, underlying architecture cause such performance gap. 
@@
 Profiling result shows that 98\% of total genome scan time is spent on memory transfer. 
-To cope with this limitation, instead of off-loading the entire correlation matrix, we use GPU to calculate the max lod score of each phenotype, and output the max.
+To cope with this limitation, instead of off-loading the entire correlation matrix, we use GPU to calculate the max lod score of each phenotype and output the max.
 
@@
 However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative \citep{CUDAnative.jl-2018} and CuArrays affords much lower programming efforts compared to C/C++.
-Compared to writting GPU functions in C/C++, writing in Julia is cleaner and easier, because it requires much less boiler plate code.
+Compared to writting GPU functions in C/C++, writing in Julia is cleaner and easier because it requires much less boiler plate code.
 Below is some code snippets to demonstrate GPU code, calling library or writing custom kernel in Julia.
@@
 
-We used 2 BXD data sets of mouse, spleen and hippocampus data sets to test our method. 
-Data from 79 BXD strains for 35556 transcripts from spleens were collected using the Affy Mouse Gene 1.0 ST array. GN accession: GN283. Marker data was available on 7321 markers;
-Hippocampus data for 93 strains (70 BXDs) summarized at the exon level for 1,236,087 probes/probe sets.
-GeneNetwork Accession Number: GN206
-Animals were generated by Drs.
-Lu, Williams, and colleagues.
-There were usually one male and one female per strain; strain averages were used as the trait of analysis.
-All animals were young adults (oldest 160 days).
-genotype probabilities were calculated using R/qtl.
+We downloaded BXD Genotype Database (GN Accession: GN600) and two sets of gene expression data, 
+UTHSC Affy MoGene 1.0 ST Spleen (GN Accession : GN283)
+and  UMUTAffy Hippocampus Exon (GN Accession: GN206) from GeneNetwork to test our method.  
+The genotype file includes 7321 markers by 198 BXD strains, the spleen consists of 79 BXD strains by 35556 transcripts, 
+and the hippocampus does 70 BXD strains by 1,236,087 probes/probe sets.  
+Genotypes were subsetted by strains, pseudo markers were added only for the hippocampus data, 
+and genotype probabilities were computed accordingly.  We thus used genotypes of 79 strains by 7321 markers for the spleen data, and 
+those of 70 by 8570 for the hippocampus data for eQTL scan.  The whole cleaning process was performed using R/qtl, R/qtl2.
 
@@
 \subsection{Benefit of using linear model}
-Our method uses linear model, and simplified the genome scan process to basic matrix operations. 
+Our method uses the linear model and simplified the genome scan process to basic matrix operations. 
 The timing of our method is shown in Table \ref{tab:speedup-table}. 
@@
 
-To cope with this limitation of GPU, we further developed \textit{Find Max LOD} method, because main interest of genome scan is to find the maximum LOD score of every phenotype. 
-This step is highly parallelizable and can utilize GPU's massive cores, at the same time, reduce the amount of data needed to be transfered back to host. 
+To cope with this limitation of GPU, we further developed \textit{Find Max LOD} method because main interest of genome scan is to find the maximum LOD score of every phenotype. 
+This step is highly parallelizable, can utilize GPU's massive cores, and, at the same time, reduce the amount of data that needs to be transfered back to host. 
 Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method for spleen data reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
@@
 We got 7 times speed up for genome scan using GPU. 
-For hippocampus datasets, which contains over 1 million phenotypes, the GPU Find Max implementation took 3.51 seconds for double precision, and 2.26 seconds for single precision. 
+For the hippocampus dataset, which contains over 1 million phenotypes, the GPU Find Max implementation took 3.51 seconds for double precision, and 2.26 seconds for single precision. 
 As a backend service to a website, this wait time is much more preferable than the CPU timing for real time interactions.
@@
 This repository contains three folders: \textit{data}, \textit{src} and \textit{test}. 
-Using Julia language and running \code{test.jl} in \textit{test} folder will run our method using either spleen data and hippocampus data in \textit{data} folder.
+Using Julia language and running \code{test.jl} in \textit{test} folder will run our method using either spleen data or hippocampus data in \textit{data} folder.
 
 This package is developed in Julia language, uses openBLAS and CUDA10.1. 
-To use this package, one needs to have an Nvidia GPU in the test machine,
+To use this package, one needs to have a Nvidia GPU in the test machine,
 and have Julia language binary, openBLAS and CUDA library installed.
@@
 It enables researchers to do real-time interaction with the data provided in GeneNetwork. 
-Such possibility can help researchers experiment with different parameters, and results in insightful findings. 
+Such possibility can help researchers experiment with different parameters and results in insightful findings. 
 LM-GPU takes advantage of GPU's massive parallel computing power and Julia's elegant syntax for quick prototyping and implementing. 

### 2019-09-23 paper.tex
Added space to two lines.

@@
 
+The matrix multiplication with large matrices is the most computationally intensive and time-consuming 
+to perform eQTL analysis.  By dividing large matrices into small block matrices, 
+one can achieve striking efficiency in matrix multiplication in conjunction with built-in BLAS in R, 
+Intel Kernel Math Library (KML) in Revolution R/Matlab, or even GPU CUDA in Matlab.  
+For instance, GOTO BLAS, only available in 64 bit R only, and KML allow the use of multiple CPU cores and provide 14-17 times 
+better performance than built-in BLAS in R.  GPU-based computation shows 10 times better performance than 
+the best algorithm for the CPU for single precision even if Matlab provides GPU CUDA but limited support to it (cite : Matrix eQTL).
+
 %% Compare matrix eQTL GridLMM \citep{runcie2018fast}, Bolt-LMM
@@
 \subsection{Acceleration Techniques}
-We used following techniques to accelerate eQTL scans. 
+We used various techniques to accelerate eQTL scan. 
 \subsubsection{Multithreaded CPU operations}
@@
 Julia provides various packages for GPU including CUDA \citep{Nickolls:2008:SPP:1365490.1365500}, OpenCL bindings. 
-Our chosen hardware for GPU is from Nvidia, which requires its propriatory library CUDA. 
-Besides the hardware requirement, CUDA is also mature and has been recongnized in the scientific computing community. 
+Our chosen hardware for GPU is from Nvidia, which requires its proprietary library CUDA. 
+Besides the hardware requirement, CUDA is also mature and has been recognized in the scientific computing community. 
 \textit{cuBLAS} \citep{cublas} library is a fast GPU implementation of the BLAS from Nvidia. 
@@
 Precision means the smallest difference between two representable numbers.
-Floating point numbers, in scientific computing, are usually stored in double presicision. 
+Floating point numbers, in scientific computing, are usually stored in double precision. 
 Double precision takes up 8 bytes in memory, while single precision takes up 4 bytes. 
@@
 However, GPU also has its own limitations. 
-To truely utilize the maximum computing power of GPU, one needs to think creatively to work around such limitations from GPU. 
-For example, during our experiemnt, memory transfer between host and device is really slow. 
+To truly utilize the maximum computing power of GPU, one needs to think creatively to work around such limitations from GPU. 
+For example, during our experiment, memory transfer between host and device is really slow. 
 Profiling result shows that 98\% of total genome scan time is spent on memory transfer. 
-To cope with this limitation, instead of off-loading the entire correlation matrix, we use GPU to calculate the max lod score of each phenotype and output the max.
+To cope with this limitation, instead of off-loading the entire correlation matrix, we use GPU to calculate the max LOD score of each phenotype and output the max.
 
@@
 However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative \citep{CUDAnative.jl-2018} and CuArrays affords much lower programming efforts compared to C/C++.
-Compared to writting GPU functions in C/C++, writing in Julia is cleaner and easier because it requires much less boiler plate code.
+Compared to writing GPU functions in C/C++, writing in Julia is cleaner and easier because it requires much less boiler plate code.
 Below is some code snippets to demonstrate GPU code, calling library or writing custom kernel in Julia.
@@
 and the hippocampus does 70 BXD strains by 1,236,087 probes/probe sets.  
-Genotypes were subsetted by strains, pseudo markers were added only for the hippocampus data, 
+Genotypes were subsetted by strains, pseudo markers were generated and added only to hippocampus data, 
 and genotype probabilities were computed accordingly.  We thus used genotypes of 79 strains by 7321 markers for the spleen data, and 
@@
 In parallel computing, Amdahl's law is used to find the theoretical maximum speedup by improving a particular part of a program.  
-For example, if a program took ten minutes for a serial processor, and a function that takes nine our of those ten minutes can be parallelized, then, the theoretical speedup, no matter how many processors are used, can not be more than ten times, because the minimum execution time of this program is one minute.
-Therefore, profiling the entire genome scan process is the prerequisit in order to begin the optimization. 
+For example, if a program takes ten minutes for a serial processor, and a function that takes nine our of those ten minutes can be parallelized, and then, the theoretical speedup, no matter how many processors are used, can not be more than ten times because the minimum execution time of this program is one minute.
+Therefore, profiling the entire genome scan process is prerequisite to the optimization. 
 Often, profiling would consider space and time complexity. 
-Our primary concern is the time taken by each function, therefore only timing information is considered in our profiling. 
-We used Julia's built-in Sampling Profiler to find our target function for GPU because it is less intrusive compared to other profiling methods.
+Our primary concern is the time taken by each function, and therefore only timing information is considered in our profiling. 
+We used Julia's built-in Sampling Profiler to find our target function for GPU because it is less intrusive compared to the other profiling methods.
 
@@
 \begin{itemize}
-	\item Calculate standadized matrix for input matrices.
-	\item Get correlation matrix
-	\item Caclulate LOD score
+	\item Calculate standardized matrices ($G^{\ast}$, $Y^{\ast}$) for input matrices ($G$, $Y$)
+	\item Get a correlation matrix ($R$) by multiplying the standardized matrices
+	\item Calculate LOD scores 
 \end{itemize}
-Our profiling result shows that step two and three takes up 90\% of time, and both involve parallelizable matrix operations. 
-So they are our candidates for GPU accleration.
-To make fair comparison for CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
+Our profiling result shows that both of the second and third steps take up 90\% of the time and involve parallelizable matrix operations.  Hence, they are our candidates for GPU acceleration.
+To make a fair comparison for CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
 The \textit{No Find Max LOD} method shows the timing when running the scan with \textit{CPU only} and \textit{CPU\&GPU}.
-As the table shows, the CPU\&GPU combination did not show any improvement over the serial version (CPU only) in terms of time. 
-In contrast, it is slower, whether using single or double precision for the data type.
+As the table shows, the CPU\&GPU combination did not show any improvement over the serial version (CPU only) in terms of the performance time.
+That is, the former is rather slower than the latter regardless of precision for the data type.
 We used a GPU profiler \textit{nvprof} \cite{nvprof}
 to investigate what is the bottleneck of GPU. 
-Result is, 98\% of time is spent on data transfer from GPU to CPU (device to host).
+The result is, 98\% of the time is spent on data transfer from GPU to CPU (device to host).
 As shown in Figure \ref{MatrixMult}, the input matrices Y' and G are small compared to the output matrix R.
 For BXD spleen dataset, the Y' and G matrix is about 40MB %TODO: double check on this number 
-, while R matrix is about 4GB. 
+while R matrix is about 4GB. 
 Data offloading is the main bottleneck for our GPU implementation.
 
-To cope with this limitation of GPU, we further developed \textit{Find Max LOD} method because main interest of genome scan is to find the maximum LOD score of every phenotype. 
-This step is highly parallelizable, can utilize GPU's massive cores, and, at the same time, reduce the amount of data that needs to be transfered back to host. 
+To cope with this limitation of GPU, we further developed \textit{Find Max LOD} method because the main interest of genome scan is to find the maximum LOD score of every trait. 
+This step is highly parallelizable, can utilize GPU's massive cores, and, at the same time, reduces the amount of data that needs to be transferred back to host. 
 Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method for spleen data reduces CPU\&GPU execution time down to 0.079 second when using double precision. 
 Compared to CPU only, which took 0.55 seconds, this GPU modification provides 7 times speedup. 
-As we mentioned earlier, since we are porting the computation to GPU which took 90\% of total execution time, according to Amdahl's Law, the theoretical maximum speed up is 10 times. 
+As we mentioned earlier, since we are porting the computation to GPU which took 90\% of the total execution time, according to Amdahl's Law, the theoretical maximum speed up is 10 times. 
 We got 7 times speed up for genome scan using GPU. 
-For the hippocampus dataset, which contains over 1 million phenotypes, the GPU Find Max implementation took 3.51 seconds for double precision, and 2.26 seconds for single precision. 
+For hippocampus data, which contains over 1 million phenotypes, the GPU Find Max implementation took 3.51 seconds for double precision, and 2.26 seconds for single precision. 
 As a backend service to a website, this wait time is much more preferable than the CPU timing for real time interactions.

### 2019-09-23 paper.tex
Removed two spaces.
Fix one typo, and add a comma.

@@
 as many genome scans as there are phenotypes.  This can be done in an
- ``embarassingly parallel'' fashion by using standard algorithms for
- QTL analysis such as those employed by R/qtl (cite).  In practice,
+``embarrassingly parallel'' fashion by using standard algorithms for
+QTL analysis, such as those employed by R/qtl (cite).  In practice,
 this is too slow, and substantial speedups are possible.  [Mention
@@
 Floating point numbers, in scientific computing, are usually stored in double precision. 
-Double precision takes up 8 bytes in memory, while single precision takes up 4 bytes. 
-In addition to the difference in storing size, the speed for calculation using single and double precisions also varies 
+Double precision floating point numbers take up 8 bytes in memory, while single precision takes up 4 bytes. 
+In addition to the difference in storage size, the speed for calculation using single and double precision also varies 
 depending on the hardware.  For example, the GPU throughput (number of floating point calculation per second, measured in FLOPS)
@@
 Based on our profiling result, the time consuming parts of our genome scan method are matrix multiplication and element-wise operations. 
-Both are amendable to GPU's heterogeneous computing architecture, since they have no data race conditions and low data dependencies. 
+Both are amenable to GPU's heterogeneous computing architecture, since they have no data race conditions and low data dependencies. 
 However, GPU also has its own limitations. 
@@
 For example, if a program takes ten minutes for a serial processor, and a function that takes nine our of those ten minutes can be parallelized, and then, the theoretical speedup, no matter how many processors are used, can not be more than ten times because the minimum execution time of this program is one minute.
-Therefore, profiling the entire genome scan process is prerequisite to the optimization. 
+Therefore, profiling the entire genome scan process is a prerequisite for the optimization. 
 Often, profiling would consider space and time complexity. 
@@
 \end{itemize}
-Our profiling result shows that both of the second and third steps take up 90\% of the time and involve parallelizable matrix operations.  Hence, they are our candidates for GPU acceleration.
-To make a fair comparison for CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
+Our profiling result shows that the second and third steps take up 90\% of the time and involve parallelizable matrix operations.  Hence, they are our candidates for GPU acceleration.
+To make a fair comparison of the CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
 The \textit{No Find Max LOD} method shows the timing when running the scan with \textit{CPU only} and \textit{CPU\&GPU}.
@@
 
-To cope with this limitation of GPU, we further developed \textit{Find Max LOD} method because the main interest of genome scan is to find the maximum LOD score of every trait. 
-This step is highly parallelizable, can utilize GPU's massive cores, and, at the same time, reduces the amount of data that needs to be transferred back to host. 
+To cope with this limitation of GPU, we further developed \textit{Find Max LOD} method, because the main interest of genome scan is to find the maximum LOD score of every phenotype. 
+This step is highly parallelizable, can utilize GPU's massive cores, and at the same time, reduces the amount of data that needs to be transferred back to host. 
 Table \ref{tab:speedup-table} shows that \textit{Find Max LOD} method for spleen data reduces CPU\&GPU execution time down to 0.079 second when using double precision. 

### 2019-09-30 paper.tex
added cuarray reference, check off some todo

@@
+% !TeX document-id = {a69d50eb-b7fd-4cc1-86b3-4b24edd99f0d}
+% !TeX TXS-program:compile = txs:///pdflatex/[--shell-escape]
+
 \documentclass[9pt,twocolumn,twoside,lineno]{gsag3jnl}
@@
 \begin{abstract}
-The BXD strains of mice are an important reference population for
+The BXD recombinant inbred strains of mice are an important reference population for
 systems biology and genetics that have been full sequenced and deeply phenotyped.
@@
 dependent on problem size and shape (n of cases, n of genotypes, n of traits).  
-Our results provide a clear path for speeding up eQTL scans using LMMs. 
+Our results provide a clear path for speeding up eQTL scans using Linear Mixed Model. 
 Our implementation is in the Julia programming language.
@@
 scores for every pair of marker and phenotypes.  With these efforts,
-we find that it is possible to perform perform real-time eQTL scans
+we find that it is possible to perform real-time eQTL scans
 for the BXD population.
@@
 There are various matrix libraries available for CPU, such as gslBLAS and OpenBLAS \citep{wang2013augem}. 
-They target different hardware or use various techniques to get highly optimal results. 
+They target different hardware or use various techniques to get optimal results. 
 The matrix multiplication in OpenBLAS is default multithreaded and therefore does not require extra coding effort to make the CPU version of matrix multiplication run in parallel.
@@
 In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
-However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative \citep{CUDAnative.jl-2018} and CuArrays affords much lower programming efforts compared to C/C++.
+However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative \citep{CUDAnative.jl-2018} and CuArrays \citep{cuarray} affords much lower programming efforts compared to C/C++.
 Compared to writing GPU functions in C/C++, writing in Julia is cleaner and easier because it requires much less boiler plate code.
@@
 \inputminted{julia}{code/gpu.jl}
-%TODO: Add reference to CUDAnative and CuArrays
 
@@
 As shown in Figure \ref{MatrixMult}, the input matrices Y' and G are small compared to the output matrix R.
-For BXD spleen dataset, the Y' and G matrix is about 40MB %TODO: double check on this number 
+For BXD spleen dataset, Y' matrix is 17 MB, G matrix is 21 MB
 while R matrix is about 4GB. 

### 2019-10-02 paper.tex
added some reference, added a paragraph about why use GPU

@@
 
-\keywords{Linear Model \\ Genome Scan \\ GPU}
+\keywords{Linear Model \\ Genome Scan \\ BXD \\ GPU}
 
@@
 \runningauthor{Hu \textit{et al.}}
+
+% This paper aims to answer three questions :
+% 1: Why BXD : reference population
+% 2: Why Speed up : GN backend and benefit other researchers
+% 3: Why GPU: Growth of CPU computing is slowing down, GPU as a new computing platform is competitive, low cost per core, and faster.
+
+% Paper Layout:
+% Abstract 
+% Intro: 
+%P1: Why BXD; 
+%P2: Why Speed Up; 
+%P3: Why GPU 
+%P4: Other works. 
+%P5: Outline of Paper
+
+% Method: 
+%P1: Intro to LM and Acceleration techniques; 
+%P2: Data Layout; 
+%P3: LM; 
+%P4: Covariates
+%P4 Acceleration Techniques;
+
+% Result: 
+%P1: Linear Model; 
+%P2: Result from using different acceleration techniques; 
+%P3: Computing platform. 
+%P4: Software Availability; 
+%P5: Data Availability
+
+% Conclusion: 
+%P1: Our final result from acceleration. 
+%P2: Limitations; 
+%P3: Future work.
 
@@
 \correspondingauthoraffiliation{1}{Corresponding author: Saunak Sen, Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis, TN. sen@uthsc.edu}
-\vspace{-34pt}% Only used for adjusting extra space in the left column of the first page
+\vspace{-23pt}% Only used for adjusting extra space in the left column of the first page
 
+%Why BXD
 \noindent The BXD mouse strains are an important set of recombinant
@@
 have been genotyped and sequenced, once a set of strains is
-phenotyped, the data can be immediately used for QTL (quantitative
-trait locus) mapping, and associational analysis with
+phenotyped, the data can be immediately used for quantitative
+trait locus (QTL) mapping, and associational analysis with
 previously-collected phenotypes.  For ``omic'' phenotypes collected
@@
 transcriptional network construction or causal mediation analyses) are
-possible.  The GeneNetwork project ({\tt http://gn2.genenetwork.org})
+possible.
+
+%Why Speed Up
+The GeneNetwork project (\url{http://gn2.genenetwork.org})
 seeks to facilitate this process by providing a searchable database of
@@
 
-%% Computational demands in omics are growing exponentially. For
-%% instance, eQTL analysis requires processing genomewide expression and
-%% marker data; this can take minutes, hours or days depending on problem
-%% size.  Scientists want to interact with the results and real-time
-%% processing would be valuable.  Our goal is to build a backend for
-%% GeneNetwork that will enable real-time genome scans for important
-%% genetic reference populations such as the BXD population.
+%Why GPU?
+Processing large data set has been a challenge for genome scan. 
+We were blessed with Moore's Law for decades, but the Central Processing Unit (CPU) is reaching to its physical limit of transistors. 
+Graphics Processing Unit (GPU), originally used as an image processing component of a computer, has shown some compelling results of accelerating computation in various fields.
+General Purpose computing Graphics Processor (GPGPU) became popular in the early 2000s because of its ability to natively handel matrix and vector operations. 
+Such power is very attractive to the scientific computing community. 
+BXD genome scan can be simplified down to matrix operations, which is quite amendable to GPU computing.
 
+%Other Works
 To perform eQTL scans in the BXD population, one has to simply perform
@@
 ``embarrassingly parallel'' fashion by using standard algorithms for
-QTL analysis, such as those employed by R/qtl (cite).  In practice,
+QTL analysis, such as those employed by R/qtl \citep{broman2003r}.  In practice,
 this is too slow, and substantial speedups are possible.  [Mention
   ideas that have been used before] For example, by using the
-Haley-Knott algorithms using genotype probabilities on batches of
+Haley-Knott algorithms \citep{haley1992simple} using genotype probabilities on batches of
 phenotypes with the same missing data pattern, instead of using the EM
-algorithm on each phenotype individually, substantial speedups are
+algorithm \citep{lander1989mapping} on each phenotype individually, substantial speedups are
 possible. This is a well-known trick and is used by R/qtl.
@@
 recombinant inbred line), then matrix multiplication can be used to
-perform Haley-Knott regression \citep{shabalin2012matrix}.
+perform Haley-Knott regression.
 
@@
 better performance than built-in BLAS in R.  GPU-based computation shows 10 times better performance than 
-the best algorithm for the CPU for single precision even if Matlab provides GPU CUDA but limited support to it (cite : Matrix eQTL).
+the best algorithm for the CPU for single precision even if Matlab provides GPU CUDA but limited support to it  \citep{shabalin2012matrix}.
 
@@
 		the expression phenotype matrix, $G$ is the matrix of
-		genotypes, and $R=Y^{\prime}G$ is the matrix of
+		genotypes, and $R=Y^{*\prime}G^{*}$ is the matrix of
 		correlations.  The LOD scores are a function of the

### 2019-10-08 paper.tex
added a paragraph for Why GPU in introduction.

@@
 
-%Why GPU?
-Processing large data set has been a challenge for genome scan. 
-We were blessed with Moore's Law for decades, but the Central Processing Unit (CPU) is reaching to its physical limit of transistors. 
-Graphics Processing Unit (GPU), originally used as an image processing component of a computer, has shown some compelling results of accelerating computation in various fields.
-General Purpose computing Graphics Processor (GPGPU) became popular in the early 2000s because of its ability to natively handel matrix and vector operations. 
-Such power is very attractive to the scientific computing community. 
-BXD genome scan can be simplified down to matrix operations, which is quite amendable to GPU computing.
-
 %Other Works
@@
 one can achieve striking efficiency in matrix multiplication in conjunction with built-in BLAS in R, 
-Intel Kernel Math Library (KML) in Revolution R/Matlab, or even GPU CUDA in Matlab.  
-For instance, GOTO BLAS, only available in 64 bit R only, and KML allow the use of multiple CPU cores and provide 14-17 times 
-better performance than built-in BLAS in R.  GPU-based computation shows 10 times better performance than 
-the best algorithm for the CPU for single precision even if Matlab provides GPU CUDA but limited support to it  \citep{shabalin2012matrix}.
+Intel Kernel Math Library (KML) in Revolution R/Matlab, or even in Matlab using CUDA.  
+%For instance, GOTO BLAS, only available in 64 bit R only, and KML allow the use of multiple CPU cores and provide 14-17 times 
+%better performance than built-in BLAS in R.  GPU-based computation shows 10 times better performance than 
+%the best algorithm for the CPU for single precision even if Matlab provides GPU CUDA but limited support to it  \citep{shabalin2012matrix}.
+
+%Why GPU?
+Processing large data set has been a challenge for genome scan. 
+We were blessed with Moore's Law for decades, but the Central Processing Unit (CPU) is reaching to its physical limit of transistors. 
+Graphics Processing Unit (GPU), originally used as an image processing component of a computer, has shown some compelling results of accelerating computation in various fields.
+General Purpose computing Graphics Processor (GPGPU) became popular in the early 2000s because of its ability to natively handel matrix and vector operations. 
+Such power is very attractive to the scientific computing community. 
+Zhang et al. \citep{zhang2015mixed} used GPU to simultaneously dissect various genetic effects with mixed linear model.
+
+%previsou work on GPU
 

### 2019-10-08 paper.tex
added some reference in introduction

@@
 Zhang et al. \citep{zhang2015mixed} used GPU to simultaneously dissect various genetic effects with mixed linear model. 
+Chapuis et all. \citep{chapuis2013graphics} utilizes GPU to offset heavy computation to deploy various ways for a more pricise detection of QTLs threashold. 
+By using GPU-backed machine learning libraries like TensorFlow, Taylor-Weiner et al.
+\citep{taylor2019scaling} reimplemented QTL mapping  and Bayesian non-negative matrix factorization, achieving greater than 200 times speedup compared to CPU-based versions.
+The ease of using such library will push method-development for genomeic research.
 

### 2019-10-09 paper.tex
added intro to acceleration techniques

@@
 
-The matrix multiplication with large matrices is the most computationally intensive and time-consuming 
-to perform eQTL analysis.  By dividing large matrices into small block matrices, 
-one can achieve striking efficiency in matrix multiplication in conjunction with built-in BLAS in R, 
-Intel Kernel Math Library (KML) in Revolution R/Matlab, or even in Matlab using CUDA.  
+%The matrix multiplication with large matrices is the most computationally intensive and time-consuming to perform eQTL analysis.  
+%By dividing large matrices into small block matrices, one can achieve striking efficiency in matrix multiplication in conjunction with built-in BLAS in R, Intel Kernel Math Library (KML) in Revolution R/Matlab, or even in Matlab using CUDA.
 %For instance, GOTO BLAS, only available in 64 bit R only, and KML allow the use of multiple CPU cores and provide 14-17 times 
@@
 %Describe genotype and phenotype data here.
+\subsection{Information about Datasets}
+
+We downloaded BXD Genotype Database (GN Accession: GN600) and two sets of gene expression data, 
+UTHSC Affy MoGene 1.0 ST Spleen (GN Accession : GN283)
+and  UMUTAffy Hippocampus Exon (GN Accession: GN206) from GeneNetwork to test our method.  
+The genotype file includes 7321 markers by 198 BXD strains, the spleen consists of 79 BXD strains by 35556 transcripts, 
+and the hippocampus does 70 BXD strains by 1,236,087 probes/probe sets.  
+Genotypes were subsetted by strains, pseudo markers were generated and added only to hippocampus data, 
+and genotype probabilities were computed accordingly.  We thus used genotypes of 79 strains by 7321 markers for the spleen data, and 
+those of 70 by 8570 for the hippocampus data for eQTL scan.  The whole cleaning process was performed using R/qtl, R/qtl2.
+
+Figure \ref{MatrixMult} shows the data layout.
+\begin{figure}[!htb]
+
+	\caption{Schematic of data and correlation calculation: $Y$ is
+		the expression phenotype matrix, $G$ is the matrix of
+		genotypes, and $R=Y^{*\prime}G^{*}$ is the matrix of
+		correlations.  The LOD scores are a function of the
+		correlation matrix.}
+	\label{MatrixMult}
+	\includegraphics[scale = .4]{figs/YGmatrix.png}
+\end{figure}
 
@@
 Corrected p-values are often appropriate, but raw p-values should be available in the supporting materials so that others may perform their own corrections.
-Figure \ref{MatrixMult} shows the data layout.
-      \begin{figure}[!htb]
-
-	\caption{Schematic of data and correlation calculation: $Y$ is
-		the expression phenotype matrix, $G$ is the matrix of
-		genotypes, and $R=Y^{*\prime}G^{*}$ is the matrix of
-		correlations.  The LOD scores are a function of the
-		correlation matrix.}
-	\label{MatrixMult}
-	\includegraphics[scale = .4]{figs/YGmatrix.png}
-\end{figure}
 
 \subsection{Acceleration Techniques}
-We used various techniques to accelerate eQTL scan. 
+While it is true that many programs can achieve tens or even hundreds of speedup by utilizing GPU, the difference needs to be examined with more care. 
+Sometimes, the reported CPU time is using single thread, and multithreaded CPU time may bring the performance gap of CPU and GPU closer than claimed. 
+Also depending on the library chozen for CPU, speed might vary depending on whether the library is optimized. 
+We believe for a fair comparison, both CPU and GPU functions should be optimized at maximum speed. 
+The following section explains our efforts of optimization on CPU and GPU functions.
+
 \subsubsection{Multithreaded CPU operations}
@@
 %Programming simplicity of interpreted languages with speed approaching compiled languages; GPU programming and linear algebra straightforward
+Although a programming language cannot really be classified as optimization technique, the choice of programming language can affect the speed of the program. 
+The ease of use is also considered, because the cost of any program comprises not only recurring expense of running time on a peice of hardware, but also the front load expense of programming effort. 
+Shorter development time is valuable because researchers can quickly prototype and experiment with different models.
 In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
@@
 \inputminted{julia}{code/gpu.jl}
-
-\subsection{Information and Availability of Datasets}
-
-We downloaded BXD Genotype Database (GN Accession: GN600) and two sets of gene expression data, 
-UTHSC Affy MoGene 1.0 ST Spleen (GN Accession : GN283)
-and  UMUTAffy Hippocampus Exon (GN Accession: GN206) from GeneNetwork to test our method.  
-The genotype file includes 7321 markers by 198 BXD strains, the spleen consists of 79 BXD strains by 35556 transcripts, 
-and the hippocampus does 70 BXD strains by 1,236,087 probes/probe sets.  
-Genotypes were subsetted by strains, pseudo markers were generated and added only to hippocampus data, 
-and genotype probabilities were computed accordingly.  We thus used genotypes of 79 strains by 7321 markers for the spleen data, and 
-those of 70 by 8570 for the hippocampus data for eQTL scan.  The whole cleaning process was performed using R/qtl, R/qtl2.
 

### 2019-10-11 paper.tex
add genome scan including covariates

@@
 \end{eqnarray*}
-where ${X}_j$ is a matrix that may include the $y-$intercept with
+where ${X}_j$ is a matrix that may include the $y$-intercept with
 or without covariate(s), and the $j$-th candidate genetic marker
@@
 correlation matrix is simply $$R=Y ^{\ast '}G^{\ast}.$$
+
+Genome scan including covariates becomes slightly different for efficient implementation.  The idea is to project genetic markers 
+and gene expressions onto the space generated by covariates and to compute the corresponding correlation matrix just as did for one-df tests.  In other words, 
+let $Z$ be a matrix of covariates including $y$-intercept.  The orthogonal projector onto the covariate space is then $P_Z= Z(Z'Z)^{-1}Z'$.  
+The markers ($G$) and gene expressions ($Y$) are now $Y_p =P_ZY$, $G_p= P_ZG$, respectively.  Standardizing them can compute the correlation matrix as shown above.
+
 %Indicate which statistical analysis has been performed and describe the method and model applied.
@@
 Sometimes, the reported CPU time is using single thread, and multithreaded CPU time may bring the performance gap of CPU and GPU closer than claimed. 
-Also depending on the library chozen for CPU, speed might vary depending on whether the library is optimized. 
+Also depending on the library chosen for CPU, speed might vary depending on whether the library is optimized. 
 We believe for a fair comparison, both CPU and GPU functions should be optimized at maximum speed. 

### 2019-10-28 paper.tex
saving changes

--- 2019-10-09 paper.tex
@@
 \citep{taylor2019scaling} reimplemented QTL mapping  and Bayesian non-negative matrix factorization, achieving greater than 200 times speedup compared to CPU-based versions.
-The ease of using such library will push method-development for genomeic research.
+The ease of using such library will push method-development for genomic research.
 
@@
 
+\section{Results}
+
+%\subsection{Benefit of using linear model}
+Our method uses the linear model and simplified the genome scan process to basic matrix operations. 
+The timing of our method is shown in Table \ref{tab:speedup-table}. 
+The execution time without finding the maximum LOD score of phenotype using double precision is 0.55 second if we only use CPU. 
+Getting the same result using the same dataset from R/qtl took about 36.5 seconds.
+
 \begin{table*}[htbp]
@@
 		\end{tabularx}
-		\footnotetext[1]{There are overhead of using GPU, such as data transfer and setting up context, the timing shown for CPU \& GPU here included all overhead for fairness.
-We ran the genome scan process 10 times and choose the median to remove randomness of each run, and the warm up time of GPU.}
+
 		\label{tab:speedup-table}
@@
 \end{table*}
-
-\section{Results}
-
-\subsection{Benefit of using linear model}
-Our method uses the linear model and simplified the genome scan process to basic matrix operations. 
-The timing of our method is shown in Table \ref{tab:speedup-table}. 
-The execution time without finding the maximum LOD score of phenotype using double precision is 0.55 second if we only use CPU. 
-Getting the same result using the same dataset from R/qtl took about 36.5 seconds.
 
@@
 To make a fair comparison of the CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
+There are overhead of using GPU, such as data transfer and setting up context, the timing shown for CPU \& GPU here included all overhead for fairness.
+We ran the genome scan process 10 times and choose the median to remove randomness of each run, and the warm up time of GPU.
 The \textit{No Find Max LOD} method shows the timing when running the scan with \textit{CPU only} and \textit{CPU\&GPU}.

### 2019-10-29 paper.tex
modified impact of using Julia section, modifed example code, and added reference to cublas example

@@
 	\label{MatrixMult}
-	\includegraphics[scale = .4]{figs/YGmatrix.png}
+	\includegraphics[scale = .4]{figs/YGmatrixStd.png}
 \end{figure}
@@
 Sometimes, the reported CPU time is using single thread, and multithreaded CPU time may bring the performance gap of CPU and GPU closer than claimed. 
-Also depending on the library chosen for CPU, speed might vary depending on whether the library is optimized. 
-We believe for a fair comparison, both CPU and GPU functions should be optimized at maximum speed. 
+Also depending on the library chosen for CPU, speed might vary depending on whether the library is optimized for such computation or hardware. 
+We believe for a fair comparison, both CPU and GPU functions should be optimized at maximum performance and count in all necesary overhead. 
 The following section explains our efforts of optimization on CPU and GPU functions.
@@
 Matrix multiplication and element-wise operation are algorithmically free of data and function dependency, so that they are amendable to GPU's parallel computing power.
-Julia provides various packages for GPU including CUDA \citep{Nickolls:2008:SPP:1365490.1365500}, OpenCL bindings. 
+Julia provides various packages for GPU including CUDA \citep{Nickolls:2008:SPP:1365490.1365500} bindings. 
 Our chosen hardware for GPU is from Nvidia, which requires its proprietary library CUDA. 
@@
 Although a programming language cannot really be classified as optimization technique, the choice of programming language can affect the speed of the program. 
-The ease of use is also considered, because the cost of any program comprises not only recurring expense of running time on a peice of hardware, but also the front load expense of programming effort. 
-Shorter development time is valuable because researchers can quickly prototype and experiment with different models.
-In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
-However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative \citep{CUDAnative.jl-2018} and CuArrays \citep{cuarray} affords much lower programming efforts compared to C/C++.
-Compared to writing GPU functions in C/C++, writing in Julia is cleaner and easier because it requires much less boiler plate code.
-Below is some code snippets to demonstrate GPU code, calling library or writing custom kernel in Julia.
-
-\inputminted{julia}{code/gpu.jl}
+Ease of use is also considered, because the cost of any program comprises not only recurring expense of running time on a peice of hardware, but also the front load expense of programming effort. 
+Shorter development time is valuable because researchers can quickly prototype and experiment with different models, but performance should also be equally important, if not more,
 
@@
 
+\subsection{Benefit of using Julia}
+In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
+However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative \citep{CUDAnative.jl-2018} and CuArrays \citep{cuarray} affords much lower programming efforts compared to C/C++.
+Compared to writing GPU functions in C/C++, writing in Julia is cleaner and easier because it requires much less boiler plate code.
+Below is some code snippets to demonstrate GPU code, calling library or writing custom kernel in Julia. 
+The first example illustrates how to call CUBLAS from Julia, second exaple shows how to write custome kernel in Julia. 
+For reason of page limitation, we will not show the corresponding C code. 
+An example of using CuBLAS with C can be found online \citep{cublas-example-code}. 
+\inputminted{julia}{code/gpu.jl}
+
 \subsection{Software Availability}

### 2019-10-29 code/gpu.jl

@@ -1,18 +1,33 @@
-# Calling GPU BLAS library
-CuArrays.CUBLAS.gemm('T', 'N', a,b);
+## Example 1:
 
-# Writing custom kernel for calculating LOD score 
-# from correlation matrix
-function lod_kernel(input, MAX,n)
+A = rand(1000,1000)
+B = rand(1000,1000)
+# Data transfer from CPU to GPU
+d_a = CuArray(A)
+d_b = CuArray(B)
+# Do matrix multiplication on GPU by calling CuBLAS library.
+d_c = CuArrays.CUBLAS.gemm('T', 'N', d_a,d_b);
+# Data Transfer from GPU to CPU
+C = collect(d_c)
+
+## Example 2:
+
+# Custom kernel to do a matrix element-wise calculation
+function log_kernel(data, MAX)
+  # calculating GPU thread ID
   i = (blockIdx().x-1) * blockDim().x + threadIdx().x
+  # Check thread ID is in bound.
   if(i < MAX+1)
-    r_square = (input[i]/n)^2
-    input[i] = (-n/Float64(2.0)) *  
-               CUDAnative.log(Float64(1.0)-r_square)
+    # Call log function on GPU
+    data[i] = CUDAnative.log(data[i])
   end
   return
 end
 
+# initialize and transfer data to GPU.
+MAX = 64000
+d_data = CuArray(rand(MAX))
 # Launching GPU
-@cuda blocks=1000 threads=64 
-lod_kernel(input, MAX, n)
+d_res = @cuda blocks=1000 threads=64 log_kernel(data, MAX)
+# Transfer result back to CPU
+res = collect(d_res)

### 2019-11-04 paper.tex
added conclusion part

@@
 
-LM-GPU provides a fast backend for GeneNetwork that is powered by GPU. 
-It enables researchers to do real-time interaction with the data provided in GeneNetwork. 
+In this paper, we experimented the effect of different matrix size on GPU speed up. 
+Matrix with closer to a square shape achieves the best speed up compared to CPU. 
+With the maximum total input data size of 16 GB, matrix multiplication is up to 5 times faster on GPU than a 16 threaded CPU. 
+LM-GPU takes advantage of GPU's massive parallel computing power and Julia's elegant syntax and fast performance for quick prototyping and performance-minded implementation in one language. 
+LM-GPU also uses vairous techniques such as multi-threaded operations, and finding the maximum lod score on GPU to acheive close to real-time genome scan.
+We are able to run genome scan for spleen data set in 0.06 seconds, and hippocampus data with over one million phenotypes in 3 seconds. 
+As a backend for GeneNetwork that is powered by GPU, LM-GPU enables researchers to do real-time interaction with the data provided in GeneNetwork. 
 Such possibility can help researchers experiment with different parameters and results in insightful findings.
-LM-GPU takes advantage of GPU's massive parallel computing power and Julia's elegant syntax for quick prototyping and implementing. 
-We are able to run genome scan for hippocampus data with over one million phenotypes in three seconds.
+
+LM-GPU package also has some limitations. 
+Currently the best speed up of GPU uses an algorithm that finds the maximum lod score to minimize data transfer between GPU and CPU, which was the bottleneck for our LM algorihtm. 
+For those who are interested in the whole correlation matrix, this approach will be a limitation. 
+LM-GPU also only supports 1-degree freedom test currently, with no missing data in input dataset. 
+Currently the package only uses linear model, but we plan to add linear mixed model in future development.
 
